---
title: "Lesson6_R_Lists"
author: "Paul Lott"
date: "2024-05-13"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Lists

A `list` in R is a data structure that is similar to `vector`, however, they can store different datatypes. Recall, `vectors` are atomic - meaning they can only store a single datatype.

Lists on the other hand can contain any different datatype or datastructure - so you can have a list containing numbers, strings, vectors, factors, lists, matrices,...

## Creating a List

You can create a list using the `list()` function.

```{r}
my_list = list(c("Hello", "World"), c(1,2,3), 1:20, TRUE, 2L)
print(my_list)
```

## Accessing Elements from a List

We can access the elements from the list using `[]` operator. But remember, that even basic datatypes are stored in R as a vector, so what you're actually getting from the list is a `vector` of your datatype.

```{r}
print(my_list[1])
```

To actually access the character string, we need to use the `[[]]` double brackets

```{r}
a = my_list[[1]]
print(a)
a[1]
```

## Changing List Elements

You can make assignment to the list or change items in the list by assigning to a position within the list.

```{r}
my_list[1] = "Goodbye"
print(my_list)
```

## Combining Lists

To combine lists, we create a vector of lists

```{r}
list1 = list("A", "B", "C")
list2 = list(1, 2 ,3)

combined_list = c(list1, list2)
print(combined_list)
```

## Converting a List to a Vector

To convert a list to a vector, we can use the `unlist()` function

```{r}
list1 = list("A", "B", "C")
list2 = list(1, "2" ,3)

print(list2)

print(unlist(list2))
```

## Named Lists

Just like Named Vectors, we can create named Lists where a value is associated with a name. But unlike `named vectors` we're not limited to a single datatype of `value`.

```{r}
# Create a list containing a vector, a matrix and a list.
list_data <- list(c("Jan","Feb","Mar"), matrix(c(3,9,5,1,-2,8), nrow = 2), list("green",12.3))

# Give names to the elements in the list.
names(list_data) <- c("1st Quarter", "A_Matrix", "A Inner list")

# Show the list.
print(list_data)
```

### Accessing Element in Named List

Like named vectors we can access by name `[]` or `[[]]` but, we can also access the items using `$` operator

```{r}
print(list_data[1])

print(list_data['1st Quarter'])

print(list_data[['1st Quarter']])

print(list_data$'1st Quarter')

print(list_data$A_Matrix)
```

Notice that the `$` operator functions similarly to `[[]]`
