---
title: "Lesson5_R_Vectors"
author: "Paul Lott"
date: "2024-05-13"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

------------------------------------------------------------------------

# R Objects - Vectors

In some of the previous examples, when we split or matched patterns in
strings - we got a `list` or `vector` out of the functions.

While `numeric`, `character`, `integer` are the base datatypes in R. We
can combine these together to create more complex data structures.

R objects combine multiple instances of a basic datatypes to create more
complex data structures.

| R Datastructure | Description                                                                                                                                                                                                    |
|-----------------|----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| vector          | A vector contains an ordered collection indexed by the integers 1, 2, …, n, where n is the length of the vector.                                                                                               |
| factor          | A factor is a special kind of vector which contains underlying numeric values 1, 2, …, n, but each of these n values has an associated character label (which may or may not be the numeric value).            |
| matrix          | A matrix contains a two-dimensional collection of numeric or character values indexed by pairs of integers (i, j).                                                                                             |
| array           | An array contains an n-dimensional collection of numeric or character values indexed by n-tuples of integers (e.g., (i,j,k) for a 3-dimensional array).                                                        |
| list            | A list contains an ordered collection of objects, and the objects in the list can be of different types.                                                                                                       |
| data.frame      | A data.frame is a type of list commonly used to store datasets, with a row for each observation and a column for each variable, and each variable can be of a different type (e.g., numeric, character, etc.). |

## R's Big Secret

While we've covered the Characters, Numerical and Logical datatypes.
We've been ignoring an important part of R that distinguishes it from
almost every other programming language.

Vectors - All of the basic datatypes are stored as a vector of length 1

```{r collapse=TRUE}
x = "Hello"
print(paste("Is a character a vector:", is.vector(x)))
print(paste("How big of a vector:", length(x)))


y = 1.02
print(paste("Is a numerical a vector:", is.vector(y)))
print(paste("How big of a vector:", length(y)))

z = TRUE
print(paste("Is a Logical a vector:", is.vector(z)))
print(paste("How big of a vector:", length(z)))

a = 8L
print(paste("Is a integer a vector:", is.vector(a)))
print(paste("How big of a vector:", length(a)))

b = raw(1)
print(paste("Is a raw a vector:", is.vector(b)))
print(paste("How big of a vector:", length(b)))

```

## Vector

Vectors is a collection of one datatype. In R, we refer this to
*atomic* - meaning that all the elements in the `vector` are going to be
stored the same `mode` in other works - the same fundamental datatype.
This means that while you can define a vector with numbers and strings -
it'll default to a single mode (ie. character) datatype.

We typically define vectors using `c(...)` notation. However, if you
want a vector with a particular mode (ie. datatype) and length, you can
create it using `vector(mode=x, length=n)`

Vectors have a very special place in R. Coming from Python, it may seem
a little weird that in R almost everything is a vector. But, yes,
Vectors are one of the fundamental data structures of R.

### Creating a Vector

```{r}
x = c(1.5)
print(x)
print(paste("MODE:", mode(x)))
print(paste("CLASS:", class(x)))
print(paste("TYPEOF:", typeof(x)))
print(paste("IS VECTOR:", is.vector(x)))
```

```{r}
x = c(1.5, 2L)
print(x)
print(paste("MODE:", mode(x)))
print(paste("CLASS:", class(x)))
print(paste("TYPEOF:", typeof(x)))
print(paste("IS VECTOR:", is.vector(x)))
```

```{r,collapse=TRUE}
x = c(1.5, "Hello", 2L)
print(x)
print(paste("MODE:", mode(x)))
print(paste("CLASS:", class(x)))
print(paste("TYPEOF:", typeof(x)))
print(paste("IS VECTOR:", is.vector(x)))
```

```{r,collapse=TRUE}
x = c(1.5, 0.223, 2L)
print(x)
print(paste("MODE:", mode(x)))
print(paste("CLASS:", class(x)))
print(paste("TYPEOF:", typeof(x)))
print(paste("IS VECTOR:", is.vector(x)))
```

```{r,collapse=TRUE}
x = c(TRUE, FALSE, TRUE)
print(x)
print(paste("MODE:", mode(x)))
print(paste("CLASS:", class(x)))
print(paste("TYPEOF:", typeof(x)))
print(paste("IS VECTOR:", is.vector(x)))
```

```{r}
x = c(TRUE, FALSE, 'TRUE')
print(x)
print(paste("MODE:", mode(x)))
print(paste("CLASS:", class(x)))
print(paste("TYPEOF:", typeof(x)))
print(paste("IS VECTOR:", is.vector(x)))
```

### Creating Vectors by Sequence

In R, there are two similar functions.

#### Colon Operator

Colon operator takes two numbers separated by a `begin:end` and creates
a `vector` that starts at begin and ends at or before `end` value. It
will only increment by 1, so if the ending value is less than one from
the last value, it'll be excluded.

If the beginning value is more than then ending value, then the values
are incremented by -1.

```{r}
x = 1:20
print(x)

(x = 1:19.9)

x = 1.1:10
print(x)

x = 10:1
print(x)

x = -1:2
print(x)
```

#### Sequence Operator

The sequence function
`seq(from, to, by, length.out=NULL, along.with=NULL)`. Allows you to
generate a sequence that increments by any value.

```{r}
print(seq(1,20))

print(seq(1,10,by=0.1))

print(seq(1,10,length.out=25))
```

### Indexing Vectors

Vectors can be indexed using the `[]`

Note: **R is 1-based.** The first element is at position 1

You can use the `:` operator, or `seq` function, or other comparison
operators to subset vectors.

```{r}
(x = seq(1,10,length.out=25))

print(x[15])

print(x[2]*3)

print(x[1:10])

print(x[c(1,3,5,25)])

print(x[x>1.5 & x<3.75])
```

Negative indices in R are not considered from the reverse end of the
vector - but rather negative values denote indices that should be
excluded.

```{r}
x = seq(1,10)

print(x[-(2:5)])
```

------------------------------------------------------------------------

## Vector Functions

We've seen functions and operators in R, for Character strings and
numerical datatypes.

### What about all the previous functions?

They still work! They'll just work on a vector of length 1 or a vector
of length 100.

However, it's important to understand how they work when you do math
with a scalar vs multiple vectors

### Scalar math on a Vector

If you are performing function on a single vector - it's as expected. R
will apply the function to each element in the vector.

```{r}
x = 1:10
print(x+12)
print(x*2)
print(cos(x))
print(x^2)
```

### Vector-Vector Math

If you use two vectors of the same length. You can think of the value of
each vector being applied to the corresponding value in the other vector

|          |     |     |     |     |     |     |     |     |     |     |
|----------|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|
| Vector 1 | 1   | 2   | 3   | 4   | 5   | 6   | 7   | 8   | 9   | 10  |
| Vector 2 | 11  | 12  | 13  | 14  | 15  | 16  | 17  | 18  | 19  | 20  |
| Addition | 12  | 14  | 16  | 18  | 20  | 22  | 24  | 26  | 28  | 30  |

```{r}
x = 1:10
y = 11:20

print(x+y)
```

If you use two vectors of differing size, then R will duplicate the
shorter of the 2 until it's the same length as the longer vector.
However, when doing numerical math if the shorter vector is not multiple
of the length of the longer vector - it'll warn you and produce an
error.

Vector 1 = c(1,2,3,4,5,6,7,8,9,10,11,12)

Vector 2 = c(11,12,13)

|          |     |     |     |     |     |     |     |     |     |     |     |     |
|----------|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|
| Vector 1 | 1   | 2   | 3   | 4   | 5   | 6   | 7   | 8   | 9   | 10  | 11  | 12  |
| Vector 2 | 11  | 12  | 13  | 11  | 12  | 13  | 11  | 12  | 13  | 11  | 12  | 13  |
| Addition | 12  | 14  | 16  | 15  | 17  | 19  | 18  | 20  | 22  | 21  | 23  | 25  |

```{r}
x = 1:12
y = c(0.1, 0.2, 0.29, 1, 4)

print(x)
print(y)
print(x+y)
```

With character string vectors of unequal length, functions such at
`paste()`- notice what happens when we `paste()` the two vectors
together.

```{r}
x = c("one", "two")
y = c(1,2,3,4,5)
paste(x,y)
```

The shorter vector is repeated, but if it doesn't exactly match up then
those values are excluded.

## Subsetting Vector

Selecting element from the vector is called Subsetting. The syntax is
shared for addition R datastructures (List, Matrices, Dataframes,
Arrays).

**Selecting multiple elements**

Use `[]` to select any number of elements from a vector. To illustrate,
I’ll apply `[]` to 1D atomic vectors, and then in the later section
about more complex objects and more dimensions we'll go over how it
works with those.

Let’s explore the different types of subsetting with a simple vector,
`x`.

```{r}
x <- c(2.1, 4.2, 3.3, 5.4, 1.5, 6.6, 9.7)
```

Note that the number after the decimal point represents the original
position in the vector.

There are six things that you can use to subset a vector:

1.  **Positive integers** return elements at the specified positions:

    ```{r}
    x[c(3, 1)] 
    ```

    Duplicate indices will duplicate values

    ```{r}
    x[c(1, 1)]
    ```

    Real numbers are silently truncated to integers

    ```{r}

    x[c(2.1, 2.9)]
    ```

2.  **Negative integers** exclude elements at the specified positions:

    ```{r}
    x[-c(3, 1)]
    ```

    Note that you can’t mix positive and negative integers in a single
    subset:

    ```{r, error=TRUE}
    x[c(-1, 2)] 
    ```

3.  **Logical vectors** select elements where the corresponding logical
    value is `TRUE`. This is probably the most useful type of subsetting
    because you can write an expression that uses a logical vector:

    ```{r}
    print(x)
    x[c(TRUE, TRUE, FALSE, FALSE)] 
    x[x > 3] 

    ```

    In `x[y]`, what happens if `x` and `y` are different lengths? The
    behaviour is controlled by the **recycling rules** where the shorter
    of the two is recycled to the length of the longer. This is
    convenient and easy to understand when one of `x` and `y` is length
    one, but I recommend avoiding recycling for other lengths because
    the rules are inconsistently applied throughout base R.

    ```{r}
    x[c(TRUE, FALSE)]  #Note differences in vector lengths x and logical (R repeats logical vector)
    ```

    Note that a missing value in the index always yields a missing value
    in the output:

    ```{r}
    print(x)
    x[c(TRUE, TRUE, NA, FALSE)]
    ```

4.  **Nothing** returns the original vector. This is not useful for 1D
    vectors, but, as you’ll see shortly, is very useful for matrices,
    data frames, and arrays. It can also be useful in conjunction with
    assignment.

    ```{r}
    x[]
    ```

5.  **Zero** returns a zero-length vector. This is not something you
    usually do on purpose, but it can be helpful for generating test
    data.

    ```{r}
    x[0]
    ```

## Additional Vector Functions

| Vector Function | Description                                |
|-----------------|--------------------------------------------|
| length(x)       | Return length of vector                    |
| sort(x)         | Sort or order a vector                     |
| order(x)        | Returns vector of indices of sorted values |
| unique(x)       | Returns only unique items                  |
| rep(x, ...)     | Replicate Elements of Vectors              |
| union(x,y)      | Returns a union of x and y                 |
| intersect(x,y)  | Returns the intersection of x and y        |

```{r}
x = c(1,2,3,4)
print(length(x))
```

```{r}
x = c(3, 2, 1, 4, 10, 23, 1, 22, 2)
print(sort(x))
print(order(x))
print(unique(x))
```

### Changing the Length of a Vector

Vectors also have some unique behavior when we want to change the length
of the vector.

However, in R - it just extends the list and will insert **NA** in the
positions where the vector was previously undefined.

```{r}
x = c(1)
x[5] = 210
print(x)
```

To shorten a vector, you can also just assign a different value to the
`length(x)` function.

```{r}
x = c(1, NA, NA, NA, 210)
length(x) = 2
print(x)
```

------------------------------------------------------------------------

# - Start 4/29/2026

## Element-wise Logical AND/OR Operators for Vectors

Remember, R loves vectors. When we use a comparison operator with a
vector the result is a vector of logicals

```{r}
x = c(1,2,3,4)
x>2
```

Element-wise AND Operator `&` evaluates for each item and returns a
logical vector.

Whereas, Logical AND operator only returns a single `TRUE` or `FALSE`

If we use `&` of `|` - we may get a logical vector back - because the
AND/OR is evaluated for each item.

```{r}
c(TRUE, TRUE) & c(TRUE, FALSE)
```

While we may not always think about writing the logical vector above.
Here is an example where we're using comparison operators to and logical
operators to select a subset of the original vector.

```{r}
x = c(3.4, 20)

print(x>2)
print(x<15)

x>2 & x<15
```

```{r}
x[x>2 | x<15]
```

Whereas, if we are using the `&&` or `||` - we'll get an error because
`&&` or `||` only want to return a single `TRUE` or `FALSE`

```{r error=TRUE}
c(TRUE, TRUE) && c(TRUE, FALSE)
```

```{r}
c(TRUE, TRUE, FALSE, FALSE) | c(TRUE, FALSE, TRUE, FALSE)
```

## Any or All

`any()` is a function that works on a logical vector and will return
`TRUE` if any values are `TRUE` and `FALSE` if none are.

`all()` is the complement function to `any()`. It will only return
`TRUE` if all the values are `TRUE`.

```{r}
any(c(TRUE, FALSE))
```

```{r}
all(c(TRUE, TRUE))
```

## %in% Operator

Checks if an element is in the vector and returns a logical vector,
where TRUE represents the position where the value are in the vector

```{r}
x = c(1,2,3,4,5,6,7,8)
y = c(2,4,6,8,10,12,14,16)

x %in% y
```

```{r}
(subset = x[x %in% y])  # Return the subset of values from X that are in Y
```

------------------------------------------------------------------------

# Named Vectors

That's because these aren't base data structures in R. In R, we can use
a `vector` similarly to a dictionary, were we assign keys as `names` and
then the value of those `names`. **But the names don't have to be
unique.** In this way, Named vectors are different Dictionaries or
Hashes.

## Creating a Named Vector

All at once, with Names:

```{r}
x = c("Lines"=10, "Yards"=2)
print(x)
```

or

In two steps,

1.  we first create the vector
2.  then associate names/keys

with the values in the vector

```{r}
x = c(10, 2, 3)
keys = c("Lines", "Yards", "Lines")
names(x) = keys

print(x)
```

Notice that we have two names that are the same.

```{r}

```

## Accessing the Value

To get the value from the array, we can then use the name/key.

We can use the indexing operator `[]` to get the name and value by
indices. Or we can use the name like we did in Python.

```{r}
print(x[3])

print(x['Lines'])
```

You'll notice that it's returning both the `Name` and `Value`. If we
want to access only the value, we need to use `[[]]` double square
brackets.

```{r}
x = c("Lines"=10, "Yards"=2, "Yards"=4)
print(x)

print(x[[2]])

print(x[['Yards']])
```

**Note:** When using Named Vectors, you're still limited by the fact
that `vectors` are atomic types. Meaning you can't mix different
datatypes as `values`. To use different datatypes, see `Named Lists` or
`hash`

## Subsetting Named Vectors

6.  If the vector is named, you can also use **character vectors** to
    return elements with matching names.

```{r}
x = c(1, 2, 3, 4)
(y <- setNames(x, c('a','b','c', 'd')))

```

Accessing value from names

```{r}
y[c("d", "c", "a")] 
```

Like integer indices, you can repeat indices

```{r}
(y[c("a", "a", "a")]) 
```

When subsetting with [, names are always matched exactly

```{r}
z <- c(abc = 1, def = 2)

(z[c("abd", "def")])
```

------------------------------------------------------------------------

# Factors

A **factor** is a vector used to specify a discrete classification
(category, grouping, or enumerated type) of the components of other
vectors. We typically use Factors in model formula. Case vs Control, or
different Categorical data.

Factors by default are considered to be ordered. However, R does support
unordered **factors**

The category/group is defined as the `level` of the factor. We can also
add separate `labels` that correspond to each `level`

## To Create a Factor

We can create a factor by starting with a vector and passing it to
`factor()`

```{r}
x = c('Fruit', 'Vegetable', 'Fruit', 'Vegetable', 'Mineral')
(y = factor(x))
```

You'll notice that the vector of 5 elements, is part of the factor, but
it has an added attribute called `levels`

The `levels` vector contains a unique set of categories and we can get
the levels of the factor by `levels()` function.

```{r}
levels(y)
```

The `factor` creation function also has additional parameters.

```{r}
# Using Labels - Maps multiple levels to two groups

x <- c("Man", "Male", "Man", "Lady", "Female")
xf <- factor(x, levels = c("Male", "Man" , "Lady",   "Female"),
                 labels = c("Male", "Male", "Female", "Female"))
print(xf)
```

# Utilizing Factors to calculate Group specific information

The `tapply` function allows you to apply a R function across a value
vector that corresponds to a the factor array.

For example, If we have a vector of number of bushels that corresponds
to the factor. We could apply the `sum()` function to get a list of
total bushels per category or group.

```{r}
bushel = c(1, 3, 1, 2, 1)
x = c('Fruit', 'Vegetable', 'Fruit', 'Vegetable', 'Mineral')
y = factor(x)
print(tapply(bushel, y, sum))
```

### Ordered Factors

The levels of factors are stored in alphabetical order, or in the order
they were specified to `factor` if they were specified explicitly.

Sometimes the levels will have a natural ordering that we want to record
and want our statistical analysis to make use of. The `ordered()`
function creates such ordered factors but is otherwise identical to
factor. For most purposes the only difference between ordered and
unordered factors is that the former are printed showing the ordering of
the levels, but the contrasts generated for them in fitting linear
models are different.

## Reading and Writing Vectors

You can write a vector using the `write()` function and read using the
`scan()` function.

#### Syntax:

```         
x = scan(file="", what=<typeof>, sep=" ") # Input 
write(x, file="", sep=" ")
```

The `what` argument should be the typeof data: `integer()`, `double()`,
`character()`, `logical()`

```{r}
x = c(1, 1, 2, 34, 2.8, 2008, 2E20)
write(x, file='numeric_vector.txt')

y = scan('numeric_vector.txt', what=double())
y
```
