---
title: "Lesson12_R_Functions"
author: "Paul Lott"
date: "2024-05-13"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Functions

Functions in R are created using the command `function()`. A function can have as many arguments as needed and is defined by assigning a variable which will be the function name `<function name>`.

You can then run the function by calling the `<function name>()`. The function code block is placed between curly brackets `{ Your code goes here }`. Finally, the values you want to return from the function must be placed in the `return(<Values you want to return goes here>)`.

#### Syntax

```         
x = function(arguments){
  <code>
  return(result)
}

y = x()  #Using your function
```

### Example

#### Reverse Complement Function

```{r}
library(stringr)
library(stringi)

dna = "ACGTT"

dna = str_to_upper(dna)
  
complement = chartr("ACGTN", "TGCAN", dna) 
rc = stri_reverse(complement)
rc
```

```{r}
library(stringr)
library(stringi)
rev_comp = function(dna, to_upper){
  # Input: 
  # - dna is DNA character sequence
  # - to_upper is a Logical (TRUE/FALSE) to change the sequence to upper case
  
  if (to_upper){
    dna = str_to_upper(dna)
  }
  
  complement = chartr("ACGTN", "TGCAN", dna) 
  rc = stri_reverse(complement)
  return(rc)
}

rev_comp("ACGTTTT", TRUE)
```

```{r}
rev_comp("GGGTTAG", TRUE)
```

## Default Arguments

Similar to Python, you can add default values for function arguments by assigning them a value. If the value is not supplied by the user then the *default* value is used.

```{r}
library(stringi)
rev_comp = function(dna, to_upper=TRUE){
 
  
  complement = chartr("acgtnACGTN", "tgcanTGCAN", dna) 
  rc = stri_reverse(complement)
  
  if (to_upper){
    rc = str_to_upper(rc)
  }
  
  return(rc)
}
```

```{r}
rev_comp("aagcagagaaag", to_upper = FALSE)
```

## Dot Arguments- Ellipsis `...` allows you to have an unlimited number of arguments

```{r}
concat = function(...){
  val = paste(...)
  return(val)
}

concat('1', '2', '4')
```

```{r}
concat("Hello", "World")
```

# Assignment #2 Create a Function from your calculate GC Percentage Code

In Lesson 4, you were assigned to take the GC percentage of a DNA sequence. For Assignment #2, convert your code to a function that takes the following:

Inputs: - DNA/RNA sequence

Arguments/Parameters:

-   remove_ambig: (Default Value: TRUE) Takes a Logical (TRUE/FALSE) whether to remove ambiguous characters before calculating GC percent

-   percentage: (Default Value: FALSE) Takes a Logical (TRUE/FALSE) whether to return value as Percentage

**Note:** You'll probably need to add some conditional blocks to test the values of your arguments, and determine how to process the sequence and return value.

```         
library(stringr) 

calc_gc = function(dna, <paramenters> ){
\# <your code> 
return(gc_perc) 
}  

calc_gc("ACGATCNNnnnnnnnnnnNNNNCCGG") would return 0.7

calc_gc("ACGATCNNnnnnnnnnnnNNNNCCGG", remove_ambig=FALSE) return 0.2692308  

calc_gc("ACGATCNNnnnnnnnnnnNNNNCCGG", remove_ambig=FALSE, percentage=TRUE) return 26.92308
```
