---
title: "Week5_Assignments"
output:
  pdf_document: default
  html_document: default
date: "2025-04-29"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment 1: Calculate the GC Percentage of DNA Sequence

For this next assignment, you'll be required to calculate the GC
percentage of a DNA sequence.

The percentage of Guanine and Cytosine in DNA or RNA significantly
impacts it's molecular stability, shape, and function.

$$
GCpercent=(\sum G +C)/\sum A + C+ G+T
$$

Using the R string functions (I'd encourage you to use the `stringr`
and/or `stringi` library), calculate the GC percentage for the following
DNA sequence. Be certain to exclude all ambiguous nucleotides (any
nucleotide that isn't ACGT).

```{r}
library(stringr)
dna = "ttcccagcacagaggaaaaggtagatctgaatgctgaacccctatatggaagaagaaaac
tgaacaaacagaaattgtcatgctctgagagccctgaggatccccaagagatgacttgga
tgacttcgaagagtagcctacagaaagttaatgattggttttctagaagtgatgatgtat
taacttctgatgatttccatgacgaagggtctaattcaaatacaaaagctgaggcggaag
aaatcccaagtgcagcagatggggtttttgtttcttcagagannnnnnnnnnnnnnnnnn
nnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnngtagaga
gtagcattgaagataaaatatttgggaaaacttatcggaggaaagcaagcttcgctaact
tgaactgcacaactgaagatgtaactctagaatcatctctactagaaccgcatatggcac
acaaacaccccttcacaaataaattaaaacgtaaaagaattacatcaagccttggtcctg
aggattttataaagaaagtagatttggcggttgttgttcaaaagtctcctgaaaagaaaa
tcgagaggctcaaccaaatggatcaaaatggtcaggtggtgaatactacta"

(dna)
```

```{r}
uc_dna = str_to_upper(dna)
cleaned_dna = str_replace_all(uc_dna, '[\n\r]', '')
cleaned_dna = str_replace_all(cleaned_dna, 'N','')
#cleaned_dna = str_replace_all(uc_dna, '[^ACGT]', '')

dna_len = nchar(cleaned_dna)
g = str_count(cleaned_dna, 'G')
c = str_count(cleaned_dna, 'C')
a = str_count(cleaned_dna, 'A')
t = str_count(cleaned_dna, 'T')

gc = g+c

gc_frac = gc/dna_len

print(gc_frac)
```

------------------------------------------------------------------------

# Assignment #2 Create a Function from from your calculate GC Percentage Code

In Lesson 4, you were assigned to take the GC percentage of a DNA
sequence. For Assignment #2, convert your code to a function that takes
the following:

Inputs: - DNA/RNA sequence

Arguments/Parameters:

-   remove_ambig: (Default Value: TRUE) Takes a Logical (TRUE/FALSE)
    whether to remove ambiguous characters before calculating GC percent

-   percentage: (Default Value: FALSE) Takes a Logical (TRUE/FALSE)
    whether to return value as Percentage

**Note:** You'll probably need to add some conditional blocks to test
the values of your arguments, and determine how to process the sequence
and return value.

```         
library(stringr) calc_gc = function(dna, ... ){ \# <your code>
return(gc_perc) }

calc_gc("ACGATCNNnnnnnnnnnnNNNNCCGG") would return 0.7

calc_gc("ACGATCNNnnnnnnnnnnNNNNCCGG", remove_ambig=FALSE) return
0.2692308

calc_gc("ACGATCNNnnnnnnnnnnNNNNCCGG", remove_ambig=FALSE,
percentage=TRUE) return 26.92308
```

```{r}
library(stringr)

calc_gc = function(dna, remove_ambig=TRUE, percentage=FALSE){
  # Input:
  # dna is a DNA string
  # 
  
  uc_dna = str_to_upper(dna)
  cleaned_dna = str_replace_all(uc_dna, '[\n\r]', '')
  
  if (remove_ambig){
    cleaned_dna = str_replace_all(cleaned_dna, '[^ACGTU]','')
  }
  
  dna_len = nchar(cleaned_dna)
  g = str_count(cleaned_dna, 'G')
  c = str_count(cleaned_dna, 'C')

  gc = g+c   # this is taking the sum of g + c

  if (percentage){
    return((gc/dna_len)*100)
  }
  
  return(gc/dna_len)
}

(calc_gc("ACGATCNNnnnnnnnnnnNNNNCCGG"))# would return 0.7

(calc_gc("ACGATCNNnnnnnnnnnnNNNNCCGG", remove_ambig=FALSE))# return 0.2692308

(calc_gc("ACGATCNNnnnnnnnnnnNNNNCCGG", remove_ambig=FALSE, percentage=TRUE)) #  return 26.92308
```
