---
title: "R Notebook"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

# PCA Analysis: Multi-dimensional Reduction

For this example, we're going to perform a PCA analysis of the
1000Genomes Dataset. We'll restrict our analysis just to Chromosome 1 in
order to make it faster, but the same analysis can be performed using
the complete dataset.

Here we'll set the environment PATH variables for the PLINK application

```{r}
Sys.setenv(PATH = paste("/cvmfs/hpc.ucdavis.edu/sw/spack/environments/main/view/generic/plink-1.9-beta7.1/bin/", Sys.getenv("PATH"), sep = ":"))
```

# Preparing the VCF File

The provided chr1.vcf.gz file is missing some IDs for variants. PLINK
uses a variant name/ID to identify markers for operations where we
remove variants.

1.  Add Variant ID (chr1:1221122:A:C) name to each variant in the VCF
2.  Import the VCF into PLINK

```{bash}
zgrep '^#' chr1.vcf.gz > chr1_vid.vcf
zgrep -v '^#' chr1.vcf.gz | awk -v OFS="\t" '{ $3=$1":"$2":"$4":"$5; print}' >> chr1_vid.vcf
plink --vcf chr1_vid.vcf --make-bed --out chr1_vid
```

# Cleaning Variants

We need to do markers and individual QC. 1. geno: Filter markers based
on missing genotyping rate \> 5% 2. mind: Filter samples with a missing
rate greater than \> 10% 3. maf: Filter variants with a variant allele
frequency less than 5% 4. nonfounders: removes individuals with at least
one parental ID 5. allow-no-sez: allows samples with no sex defined in
the BIM file

```{bash}
plink --bfile chr1_vid --geno 0.05 --mind 0.1 --maf 0.05 --nonfounders --allow-no-sex --make-bed --out chr1_cleaned
```

#LD Pruning Linkage disequilibrium (LD) pruning is a preprocessing step
in genetic analyses where highly correlated variants (typically SNPs
that are inherited together due to physical proximity on a chromosome)
are systematically removed, leaving a subset of approximately
independent markers. This is particularly important prior to performing
principal component analysis (PCA), because PCA assumes that variables
are not overly redundant. If dense clusters of correlated SNPs are
retained, they can disproportionately influence the principal
components, causing the analysis to reflect local genomic structure
(such as long haplotype blocks) rather than true genome-wide patterns of
variation. By pruning SNPs in high LD, the resulting dataset better
represents independent genetic signals, allowing PCA to more accurately
capture population structure, ancestry differences, and other
broad-scale patterns without bias from correlated regions.

Plink performs the pruning in 2 steps:

-   First, Create a list of variants to keep or remove based on
    R-squared value
-   Second, Extract the variants you want to keep for the PCA analysis.

```{bash}
plink --bfile chr1_cleaned --indep-pairwise 50 5 0.2 --out pruning_results
plink --bfile chr1_cleaned --extract pruning_results.prune.in --make-bed --out chr1_pruned
```

# Calculating Distance between all Samples

Calculate the Hamming Distance or Identity-by-State Matrix between all
samples. This leads to a large matrix that is 2318 rows x 2318 columns.
This matrix is too complex to plot in a simple manner, so we need to
reduce it.

```{bash}
plink --bfile chr1_pruned --distance-matrix --out dataForPCA
```

# PCA Analysis

## Import the datasets

```{r}
## Load data
dist_populations<-read.table("dataForPCA.mdist",header=F)
### Extract breed names
fam <- data.frame(famids=read.table("dataForPCA.mdist.id")[,1])
### Extract individual names 
famInd <- data.frame(IID=read.table("dataForPCA.mdist.id")[,2])
```

## Perform PCA

Principal component analysis (PCA) is a method for simplifying complex
datasets by transforming many correlated variables into a smaller set of
new variables—called principal components—that capture the main patterns
of variation in the data. You can think of it as finding the directions
in which the data varies the most, allowing you to summarize and
visualize high-dimensional data in just a few dimensions (often two or
three). In R, the cmdscale function performs a closely related technique
called classical multidimensional scaling (MDS), which can be used to
obtain PCA-like results when applied to an appropriate distance matrix.
Instead of working directly on the raw data matrix, cmdscale starts with
a matrix of pairwise distances between samples (for example, genetic
distances), then uses eigen decomposition to project those distances
into a lower-dimensional space. The resulting coordinates represent each
sample in terms of principal axes that approximate the original distance
relationships, effectively capturing the major structure in the data
similar to PCA.

```{r}
## Perform PCA using the cmdscale function 
# Time intensive step - takes a few minutes
mds_populations <- cmdscale(dist_populations,eig=T,5)
```

## Create DataFrames with IDs

```{r}
## Extract the eigen vectors
eigenvec_populations <- cbind(fam,famInd,mds_populations$points)
```

## Get Proportion of variation captured by each eigenvector

```{r}
## Proportion of variation captured by each eigen vector
eigen_percent <- round(((mds_populations$eig)/sum(mds_populations$eig))*100,2)
```

# Plotting PCA Analysis Results

```{r}
# Visualize PCA in tidyverse
# Load tidyverse
if (!require("tidyverse")) {
  install.packages("tidyverse", dependencies = TRUE)
  library(tidyverse)
}

if (!require("plotly")) {
  install.packages("plotly", dependencies = TRUE)
  library(plotly)
}

# PCA plot
p = ggplot(data = eigenvec_populations) +
 geom_point(mapping = aes(x = `1`, y = `2`,color = famids), show.legend = FALSE ) + 
 geom_hline(yintercept = 0, linetype="dotted") + 
 geom_vline(xintercept = 0, linetype="dotted") +
 labs(title = "PCA of wordwide goat populations",
    x = paste0("Principal component 1 (",eigen_percent[1]," %)"),
    y = paste0("Principal component 2 (",eigen_percent[2]," %)")) + 
 theme_minimal()

ggplotly(p)
```
