---
title: "Lesson13 R Plotting"
author: "Paul Lott"
date: "2024-05-13"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Plotting

I'm not going to cover R basic plots in-depth because they are very basic. We'll try to spend a little more time on `ggplot` in the Tidyverse Lessons. `ggplot` has become the defacto graphics library for creating publication plots.

Basic R provides the following types of plots:

-   Scatterplot: `plot()`

    -   Pairs: `pairs()`

-   Barplot: `barplot()`

-   Histogram: `hist()`

-   Boxplot: `boxplot()`

-   Pie Chart: `pie()`

## Scatterplot

Scatter plots can plot a single vector *(x,x)* or 2 vectors *(x,y) on a X-Y Coordinate* plot. You can specify types of *points (`pch`*) or lines (`lty`)

## Syntax

```         
plot(x, y = NULL, type = "p",  xlim = NULL, ylim = NULL,
     log = "", main = NULL, sub = NULL, xlab = NULL, ylab = NULL,
     ann = par("ann"), axes = TRUE, frame.plot = axes,
     panel.first = NULL, panel.last = NULL, asp = NA,
     xgap.axis = NA, ygap.axis = NA,
     ...)
```

+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Argument                               | Description                                                                                                                                                                                                                                                                                                                                                                       |
+========================================+===================================================================================================================================================================================================================================================================================================================================================================================+
| x, y                                   | the x and y arguments provide the x and y coordinates for the plot. Any reasonable way of defining the coordinates is acceptable. See the function xy.coords for details. If supplied separately, they must be of the same length.                                                                                                                                                |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| type                                   | 1-character string giving the type of plot desired. The following values are possible, for details, see plot: "p" for points, "l" for lines, "b" for both points and lines, "c" for empty points joined by lines, "o" for overplotted points and lines, "s" and "S" for stair steps and "h" for histogram-like vertical lines. Finally, "n" does not produce any points or lines. |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| xlim                                   | the x limits (x1, x2) of the plot. Note that x1 \> x2 is allowed and leads to a ‘reversed axis’.                                                                                                                                                                                                                                                                                  |
|                                        |                                                                                                                                                                                                                                                                                                                                                                                   |
|                                        | The default value, NULL, indicates that the range of the finite values to be plotted should be used.                                                                                                                                                                                                                                                                              |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| ylim                                   | the y limits of the plot.                                                                                                                                                                                                                                                                                                                                                         |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| log                                    | a character string which contains "x" if the x axis is to be logarithmic, "y" if the y axis is to be logarithmic and "xy" or "yx" if both axes are to be logarithmic.                                                                                                                                                                                                             |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| main                                   | a main title for the plot, see also title.                                                                                                                                                                                                                                                                                                                                        |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| sub                                    | a sub title for the plot.                                                                                                                                                                                                                                                                                                                                                         |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| xlab                                   | a label for the x axis, defaults to a description of x.                                                                                                                                                                                                                                                                                                                           |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| ylab                                   | a label for the y axis, defaults to a description of y.                                                                                                                                                                                                                                                                                                                           |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| ann                                    | a logical value indicating whether the default annotation (title and x and y axis labels) should appear on the plot.                                                                                                                                                                                                                                                              |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| axes                                   | a logical value indicating whether both axes should be drawn on the plot. Use graphical parameter "xaxt" or "yaxt" to suppress just one of the axes.                                                                                                                                                                                                                              |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| frame.plot                             | a logical indicating whether a box should be drawn around the plot.                                                                                                                                                                                                                                                                                                               |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| panel.first                            | an ‘expression’ to be evaluated after the plot axes are set up but before any plotting takes place. This can be useful for drawing background grids or scatterplot smooths. Note that this works by lazy evaluation: passing this argument from other plot methods may well not work since it may be evaluated too early.                                                         |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| panel.last                             | an expression to be evaluated after plotting has taken place but before the axes, title and box are added. See the comments about panel.first.                                                                                                                                                                                                                                    |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| asp                                    | the y/x aspect ratio, see plot.window.                                                                                                                                                                                                                                                                                                                                            |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| xgap.axis, ygap.axis                   | the x/y axis gap factors, passed as gap.axis to the two axis() calls (when axes is true, as per default).                                                                                                                                                                                                                                                                         |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| col                                    | The colors for lines and points. Multiple colors can be specified so that each point can be given its own color. If there are fewer colors than points they are recycled in the standard fashion. Lines will all be plotted in the first colour specified.                                                                                                                        |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| bg                                     | a vector of background colors for open plot symbols, see [`points`](http://localhost:8090/help/library/graphics/help/points). Note: this is **not** the same setting as [`par`](http://localhost:8090/help/library/graphics/help/par)`("bg")`.                                                                                                                                    |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| pch                                    | a vector of plotting characters or symbols: see [`points`](http://localhost:8090/help/library/graphics/help/points)                                                                                                                                                                                                                                                               |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| cex                                    | a numerical vector giving the amount by which plotting characters and symbols should be scaled relative to the default. This works as a multiple of [`par`](http://localhost:8090/help/library/graphics/help/par)`("cex")`. `NULL` and `NA` are equivalent to `1.0`. Note that this does not affect annotation: see below.                                                        |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| lty                                    | a vector of line types, see [`par`](http://localhost:8090/help/library/graphics/help/par).                                                                                                                                                                                                                                                                                        |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `cex.main`, `col.lab`, `font.sub`, etc | settings for main- and sub-title and axis annotation, see [`title`](http://localhost:8090/help/library/graphics/help/title) and [`par`](http://localhost:8090/help/library/graphics/help/par).                                                                                                                                                                                    |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| lwd                                    | a vector of line widths, see [`par`](http://localhost:8090/help/library/graphics/help/par)                                                                                                                                                                                                                                                                                        |
+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

```{r}
x = read.table("https://raw.githubusercontent.com/mwaskom/seaborn-data/master/penguins.csv", header=TRUE, sep=',')
plot(x$bill_length_mm, x$bill_depth_mm, xlab="Bill Length (mm)", ylab="Bill Depth (mm)")
```

## Pair Plot

Pair plot allow you to create scatterplots to show the relationship between 2 or more variables in your matrix or data.frame.

### Syntax

```         
pairs(~formula, data)
```

-   **\~formula** is a mathematical formula using column names, such as `~bill_length_mm+bill_depth_mm`

-   **data** is the dataset where the formula defines the column names

See: `help(pairs)` for more information

```{r}
pairs(~bill_length_mm+bill_depth_mm+body_mass_g+flipper_length_mm, data=x)
```

## Barplot

Bar plot represents a vector value as the height of the bar.

### Syntax

```         
barplot(height, width = 1, space = NULL,
        names.arg = NULL, legend.text = NULL, beside = FALSE,
        horiz = FALSE, density = NULL, angle = 45,
        col = NULL, border = par("fg"),
        main = NULL, sub = NULL, xlab = NULL, ylab = NULL,
        xlim = NULL, ylim = NULL, xpd = TRUE, log = "",
        axes = TRUE, axisnames = TRUE,
        cex.axis = par("cex.axis"), cex.names = par("cex.axis"),
        inside = TRUE, plot = TRUE, axis.lty = 0, offset = 0,
        add = FALSE, ann = !add && par("ann"), args.legend = NULL, ...)
```

-   **height** is the vector with values for height

-   **xlab, ylab** are the labels for the x and y axes

-   **main** is the title for the plot

See: `help(barplot)` or `example(barplot`) for more information on barplots

```{r}
barplot(c(1,2,3,4,5), ylab="Counts", xlab="Numbers", main ="Example Barplot")
```

## Histogram

### Syntax

```         
hist(x, breaks = "Sturges",
     freq = NULL, probability = !freq,
     include.lowest = TRUE, right = TRUE,
     density = NULL, angle = 45, col = "lightgray", border = NULL,
     main = paste("Histogram of" , xname),
     xlim = range(breaks), ylim = NULL,
     xlab = xname, ylab,
     axes = TRUE, plot = TRUE, labels = FALSE,
     nclass = NULL, warn.unused = TRUE, ...)
```

See: `help(hist)` or `example(hist`) for more information on histograms

```{r}
hist(x$flipper_length_mm, xlab="Flipper Length (mm)", main="Histogram of Penguin Flipper Length")
```

## Boxplot

### Syntax

```         
## S3 method for class 'formula'
boxplot(formula, data = NULL, ..., subset, na.action = NULL,
        xlab = mklab(y_var = horizontal),
        ylab = mklab(y_var =!horizontal),
        add = FALSE, ann = !add, horizontal = FALSE,
        drop = FALSE, sep = ".", lex.order = FALSE)

## Default S3 method:
boxplot(x, ..., range = 1.5, width = NULL, varwidth = FALSE,
        notch = FALSE, outline = TRUE, names, plot = TRUE,
        border = par("fg"), col = "lightgray", log = "",
        pars = list(boxwex = 0.8, staplewex = 0.5, outwex = 0.5),
         ann = !add, horizontal = FALSE, add = FALSE, at = NULL)
```

See: `help(boxplot)` or `example(boxplot`) for more information on boxplots

```{r}
boxplot(x$flipper_length_mm, ylab="Flipper Length (mm)", main="Boxplot of Penguin Flipper Length")
```

Using the formula, we can specify a numerical vector (`x$flipper_length_mm`) that we want to split up by a categorical group (`x$species`)

```{r}
boxplot(x$flipper_length_mm ~ x$species, ylab="Flipper Length (mm)", xlab="Species", main="Boxplot of Penguin Flipper Length")
```

## Pie Chart

Circular chart split into segments based on the ratio of data provided

### Syntax

```         
pie(x, labels = names(x), edges = 200, radius = 0.8,
    clockwise = FALSE, init.angle = if(clockwise) 90 else 0,
    density = NULL, angle = 45, col = NULL, border = NULL,
    lty = NULL, main = NULL, ...)
```

See: `help(pie)` or `example(pie`) for more information on histograms

```{r}
pie(c(1,2,3,4,5,6,7), labels=c('One', 'Two', 'Three', 'Four', 'Five', 'Six', 'Seven'), main='Example Pie chart')
```

## Saving Plots

Plots by default are output to the current window or Notebook, however, if you want to save them to a file you need to specify which *graphical* device R should send the plot to prior to plotting. Although, you can export them in RStudio from the *Plots* pane. The *graphical device* are specific to the file format you want to output your plot.

| Graphical Devices | Description                        |
|-------------------|------------------------------------|
| pdf()             | Write to PDF file                  |
| postscript()      | Writes to a PostScript File        |
| svg()             | Write to a Scalable Vector Graphic |
| png()             | Write to PNG File                  |
| jpeg()            | Write to JPEG File                 |
| tiff()            | Write to TIFF File                 |
| bmp()             | Write to Bitmap File               |

After plotting, you need to turn the *graphical device* off to complete writing to the file by issuing the `dev.off()` command .

```{r}
png(file='Test.png')
plot(seq(1,2,0.1))
dev.off()
```

If your forget to turn the `graphical device` off between plotting functions, the second plot will overwrite the first plot

```{r}
png(file='Test.png')
plot(seq(1,2,0.1))
plot(2:30) # Overwrites the previous plot
dev.off()
```
