---
title: "Lesson 21: Salmon and tximeta"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

# Quantifying with *Salmon*

As mentioned above, a short tutorial on how to use *Salmon* can be found
[here](https://combine-lab.github.io/salmon/getting_started/), so
instead we will provide the code that was used to quantify the files
used in this workflow. *Salmon* can be conveniently run on a cluster
using the [Snakemake](https://snakemake.readthedocs.io/en/stable/)
workflow management system [@snakemake].

The following `Snakemake` file was used to quantify the eight samples
that were downloaded from the SRA (the SRR identifier is the run
identifier, and there was only one run per sample for these eight
samples).

```         
DATASETS = ["SRR1039508",
            "SRR1039509",
            "SRR1039512",
            "SRR1039513",
            "SRR1039516",
            "SRR1039517",
            "SRR1039520",
            "SRR1039521"]

SALMON = "/path/to/salmon_0.14.1/bin/salmon"

rule all:
  input: expand("quants/{dataset}/quant.sf", dataset=DATASETS)

rule salmon_quant:
    input:
        r1 = "fastq/{sample}_1.fastq.gz",
        r2 = "fastq/{sample}_2.fastq.gz",
        index = "/path/to/gencode.v29_salmon_0.14.1"
    output:
        "quants/{sample}/quant.sf"
    params:
        dir = "quants/{sample}"
    shell:
        "{SALMON} quant -i {input.index} -l A -p 6 --validateMappings \
         --gcBias --numGibbsSamples 20 -o {params.dir} \
         -1 {input.r1} -2 {input.r2}"
```

The last line is the key one which runs *Salmon*. It says to quantify
using a specific *index*, with automatic library type detection, using 6
threads, with the validate mappings setting (this is default in versions
of *Salmon* $\ge$ 0.99), with GC bias correction, and writing out 20
Gibbs samples (this is optional). The last three arguments specify the
output directory and the two paired read files.

The above Snakemake file requires that an index be created at
`/path/to/gencode.vVV_salmon_X.Y.Z`, where VV and X,Y,Z should help
specify the release of the reference transcripts and of *Salmon*. For
human and mouse reference transcripts, we recommend to use
[GENCODE](https://gencodegenes.org) [@gencode].

The *Salmon* index can be created easily with the following command:

```         
salmon index -t transcripts.fa.gz -i name_of_index
```

**Note:** *Salmon* can also make use of genomic *decoy sequences* during
indexing, as described in @Srivastava2020, which has been demonstrated
to improve quantification accuracy. For further details on how to make
use of decoy sequences within the *Salmon* software please consult this
note in the [Salmon
documentation](https://salmon.readthedocs.io/en/latest/salmon.html#preparing-transcriptome-indices-mapping-based-mode).
Here, we will continue mapping reads to the transcriptome without decoy
sequences. During indexing, a note will appear that the *Salmon* index
is being built without any decoy sequences.

If the transcripts are downloaded from GENCODE, it is recommended to use
something similar to the following command (which simply helps to strip
the extra information from the transcript names):

```         
salmon index --gencode -t gencode.v29.transcripts.fa.gz \
  -i gencode.v29_salmon_X.Y.Z
```

The above Snakemake file can then be used to [execute
Snakemake](https://snakemake.readthedocs.io/en/stable/executable.html)
in various ways, including submitting multiple jobs to a compute cluster
or in the cloud. The above Snakemake file was executed on a cluster with
SLURM scheduling, with the following line in a separate job submitted to
the cluster:

```         
snakemake -j 4 --latency-wait 30 --cluster "sbatch -N 1 -n 6"
```

```{r loadairway}
library("airway")
```

The R function *system.file* can be used to find out where on your
computer the files from a package have been installed. Here we ask for
the full path to the `extdata` directory, where R packages store
external data, that is part of the `r Biocexptpkg("airway")` package.

```{r dir}
dir <- system.file("extdata", package="airway", mustWork=TRUE)
```

In this directory, we find a number of files, including eight BAM files
that were used in the previous version of this workflow demonstrating
alignment and counting. We will focus on the two directories that are in
the `quants` directory, which contain the output from *Salmon* on two
files.

```{r list.files}
list.files(dir)
list.files(file.path(dir, "quants"))
```

Typically, we have a table with detailed information for each of our
samples that links samples to the associated FASTQ and *Salmon*
directories. For your own project, you might create such a
comma-separated value (CSV) file using a text editor or spreadsheet
software such as Excel.

We load such a CSV file with *read.csv*:

```{r sampleinfo}
csvfile <- file.path(dir, "sample_table.csv")
coldata <- read.csv(csvfile, row.names=1, stringsAsFactors=FALSE)
coldata
```

To demonstrate loading *Salmon* quantifiation data into R, we will just
work with the two samples that are provided in the *airway* package. We
create a column called `names` and a column called `files`:

```{r makecoldata}
coldata <- coldata[1:2,]
coldata$names <- coldata$Run
coldata$files <- file.path(dir, "quants", coldata$names, "quant.sf.gz")
file.exists(coldata$files)
```

Now we load the *tximeta* package and run its main function:

```{r tximeta, message=TRUE}
library("tximeta")
se <- tximeta(coldata)
```

If the reference transcriptome checksum was recognized by tximeta
(details on this in the `r Biocpkg("tximeta")` vignette), and if we have
a working internet connection, *tximeta* will locate and download the
relevant annotation data from various sources. A few details: the
annotation data is only downloaded and parsed once, subsequently it will
used locally cached versions of the metadata as needed (if you load data
a second time that was quantified against the same reference
transcripts). Also, the very first time that one uses *tximeta*, it will
ask you to approve the default cache location (following the paradigm of
the cache location used by other R and Bioconductor packages). You can
change this location at any point later.

We will discuss what is the structure of the `se` object in the next
section, but we can first just consider the dimensions. Note that
*tximeta* imports data at the *transcript level*.

```{r lookse}
dim(se)
head(rownames(se))
```

As this workflow is concerned with gene-level analysis, we will now
summarize the transcript-level quantifications to the gene level (which
internally makes use of the methods in *tximport*
[@Soneson2015Differential]). The correct transcript-to-gene mapping
table is automatically created based on the metadata stored within the
`se` object.

```{r summarize, message=TRUE}
gse <- summarizeToGene(se)
```

Now we can check that the dimensions are reduced and the row IDs are now
gene IDs:

```{r lookgse}
dim(gse)
head(rownames(gse))
```

## *DESeq2* import functions

While the above section described use of *Salmon* and *tximeta*, there
are many possible inputs to *DESeq2*, each of which have their own
dedicated import functions. The following tools can be used generate or
compile count data for use with *DESeq2*: *tximport*
[@Soneson2015Differential], *tximeta* [@Love2020], *htseq-count*
[@Anders2015HTSeqa], *featureCounts* [@Liao2014FeatureCounts],
*summarizeOverlaps* [@Lawrence2013Software].

| function | package | framework | output | *DESeq2* input function |
|---------------|---------------|---------------|---------------|---------------|
| *tximport* | `r Biocpkg("tximport")` | R/Bioconductor | list of matrices | *DESeqDataSetFromTximport* |
| *tximeta* | `r Biocpkg("tximeta")` | R/Bioconductor | *SummarizedExperiment* | *DESeqDataSet* |
| *htseq-count* | [HTSeq](http://www-huber.embl.de/users/anders/HTSeq) | Python | files | *DESeqDataSetFromHTSeq* |
| *featureCounts* | `r Biocpkg("Rsubread")` | R/Bioconductor | matrix | *DESeqDataSetFromMatrix* |
| *summarizeOverlaps* | `r Biocpkg("GenomicAlignments")` | R/Bioconductor | *SummarizedExperiment* | *DESeqDataSet* |

We will next describe the class of object created by *tximeta* which was
saved above as `se` and `gse`, and how to create a *DESeqDataSet* object
from this for use with *DESeq2* (the other functions above also create a
*DESeqDataSet*).

## SummarizedExperiment

```{r sumexp, echo=FALSE}
par(mar=c(0,0,0,0))
plot(1,1,xlim=c(0,100),ylim=c(0,100),bty="n",
     type="n",xlab="",ylab="",xaxt="n",yaxt="n")
polygon(c(45,90,90,45),c(5,5,70,70),col="pink",border=NA)
polygon(c(45,90,90,45),c(68,68,70,70),col="pink3",border=NA)
text(67.5,40,"assay(s)")
text(67.5,35,'e.g. "counts", ...')
polygon(c(10,40,40,10),c(5,5,70,70),col="skyblue",border=NA)
polygon(c(10,40,40,10),c(68,68,70,70),col="skyblue3",border=NA)
text(25,40,"rowRanges")
polygon(c(45,90,90,45),c(75,75,95,95),col="palegreen",border=NA)
polygon(c(45,47,47,45),c(75,75,95,95),col="palegreen3",border=NA)
text(67.5,85,"colData")
```

**The component parts of a *SummarizedExperiment* object.** The `assay`
(pink block) contains the matrix of counts, the `rowRanges` (blue block)
contains information about the genomic ranges and the `colData` (green
block) contains information about the samples. The highlighted line in
each block represents the first row (note that the first row of
`colData` lines up with the first column of the `assay`).

The *SummarizedExperiment* container is diagrammed in the Figure above
and discussed in the latest Bioconductor paper
[@Huber2015Orchestrating]. In our case, *tximeta* has created an object
`gse` with three matrices: "counts" - the estimated fragment counts for
each gene and sample, "abundance" - the estimated transcript abundances
in TPM, and "length" - the effective gene lengths which include changes
in length due to biases as well as due to transcript usage. The names of
the assays can be examined with *assayNames*, and the assays themselves
are stored as `assays` (a list of matrices). The first matrix in the
list can be pulled out via `assay`. The `rowRanges` for our object is
the *GRanges* of the genes (from the left-most position of all the
transcripts to the right-most position of all the transcripts). The
component parts of the *SummarizedExperiment* are accessed with an R
function of the same name: `assay` (or `assays`), `rowRanges` and
`colData`.

We now will load the full count matrix corresponding to all samples and
all data, which is provided in the *airway* package, and will continue
the analysis with the full data object. We can investigate this
*SummarizedExperiment* object by looking at the matrices in the `assays`
slot, the phenotypic data about the samples in `colData` slot, and the
data about the genes in the `rowRanges` slot.

```{r loadfullgse}
data(gse)
gse
```

The counts are the first matrix, so we can examine them with just
`assay`:

```{r assaysgse}
assayNames(gse)
head(assay(gse), 3)
colSums(assay(gse))
```

The `rowRanges`, when printed, shows the ranges for the first five and
last five genes:

```{r rowrangesgse}
rowRanges(gse)
```

The `rowRanges` also contains metadata about the sequences (chromosomes
in our case) in the `seqinfo` slot:

```{r lookseqinfo}
seqinfo(rowRanges(gse))
```

The `colData` for the *SummarizedExperiment* reflects the *data.frame*
that was provided to the `tximeta` function for importing the
quantification data. Here we can see that there are columns indicating
sample names, as well as the donor ID, and the treatment condition
(treated with dexamethasone or untreated).

```{r coldatagse}
colData(gse)
```

## Branching point

# The *DESeqDataSet* object, sample information and the design formula

Bioconductor software packages often define and use a custom class for
storing data that makes sure that all the needed data slots are
consistently provided and fulfill the requirements. In addition,
Bioconductor has general data classes (such as the
*SummarizedExperiment*) that can be used to move data between packages.
Additionally, the core Bioconductor classes provide useful
functionality: for example, subsetting or reordering the rows or columns
of a *SummarizedExperiment* automatically subsets or reorders the
associated *rowRanges* and *colData*, which can help to prevent
accidental sample swaps that would otherwise lead to spurious results.
With *SummarizedExperiment* this is all taken care of behind the scenes.

In *DESeq2*, the custom class is called *DESeqDataSet*. It is built on
top of the *SummarizedExperiment* class, and it is easy to convert
*SummarizedExperiment* objects into *DESeqDataSet* objects, which we
show below. One of the two main differences is that the `assay` slot is
instead accessed using the *counts* accessor function, and the
*DESeqDataSet* class enforces that the values in this matrix are
non-negative integers.

A second difference is that the *DESeqDataSet* has an associated *design
formula*. The experimental design is specified at the beginning of the
analysis, as it will inform many of the *DESeq2* functions how to treat
the samples in the analysis (one exception is the size factor
estimation, i.e., the adjustment for differing library sizes, which does
not depend on the design formula). The design formula tells which
columns in the sample information table (`colData`) specify the
experimental design and how these factors should be used in the
analysis.

First, let's examine the columns of the `colData` of `gse`. We can see
each of the columns just using the `$` directly on the
*SummarizedExperiment* or *DESeqDataSet*.

```{r gsevars}
gse$donor
gse$condition
```

We can rename our variables if we want. Let's use `cell` to denote the
donor cell line, and `dex` to denote the treatment condition.

```{r gsevarsrename}
gse$cellLine <- gse$donor
gse$dexamethasone <- gse$condition
```

We can also change the names of the levels. It is critical when one
renames levels to not change the order. Here we will rename
`"Untreated"` as `"untrt"` and `"Dexamethasone"` as `"trt"`:

```{r renamelevels}
levels(gse$dexamethasone)
# when renaming levels, the order must be preserved!
levels(gse$dexamethasone) <- c("untreated", "treated")
```

The simplest design formula for differential expression would be
`~ condition`, where `condition` is a column in `colData(dds)` that
specifies which of two (or more groups) the samples belong to. For the
airway experiment, we will specify `~ cell + dex` meaning that we want
to test for the effect of dexamethasone (`dex`) controlling for the
effect of different cell line (`cell`).

**Note:** it is prefered in R that the first level of a factor be the
reference level (e.g. control, or untreated samples). In this case, when
the `colData` table was assembled the untreated samples were already set
as the reference, but if this were not the case we could use *relevel*
as shown below. While `levels(...) <-` above was simply for renaming the
character strings associated with levels, *relevel* is a very different
function, which decides how the variables will be coded, and how
contrasts will be computed. For a two-group comparison, the use of
*relevel* to change the reference level would flip the sign of a
coefficient associated with a contrast between the two groups.

```{r gsedex}
library("magrittr")
gse$dexamethasone %<>% relevel("untreated")
gse$dexamethasone
```

`%<>%` is the compound assignment pipe-operator from the
`r CRANpkg("magrittr")` package, the above line of code is a concise way
of saying:

```{r explaincmpass, eval = FALSE}
gse$dexamethasone <- relevel(gse$dexamethasone, "untreated")
```

For running *DESeq2* models, you can use R's formula notation to express
any fixed-effects experimental design. Note that *DESeq2* uses the same
formula notation as, for instance, the *lm* function of base R. If the
research aim is to determine for which genes the effect of treatment is
different across groups, then interaction terms can be included and
tested using a design such as `~ group + treatment + group:treatment`.
See the manual page for `?results` for more examples. We will show how
to use an interaction term to test for condition-specific changes over
time in a time course example below.

In the following sections, we will demonstrate the construction of the
*DESeqDataSet* from two starting points:

-   from a *SummarizedExperiment* object
-   from a count matrix and a sample information table

For a full example of using the *HTSeq* Python package for read
counting, please see the `r Biocexptpkg("pasilla")` vignette. For an
example of generating the *DESeqDataSet* from files produced by
*htseq-count*, please see the `r Biocpkg("DESeq2")` vignette.

## Starting from *SummarizedExperiment*

Again, we can quickly check the millions of fragments that could be
mapped by *Salmon* to the genes (the second argument of *round* tells
how many decimal points to keep).

```{r countreads}
round( colSums(assay(gse)) / 1e6, 1 )
```

Once we have our fully annotated *SummarizedExperiment* object, we can
construct a *DESeqDataSet* object from it that will then form the
starting point of the analysis. We add an appropriate design for the
analysis:

```{r loaddeseq2}
library("DESeq2")
```

```{r makedds}
dds <- DESeqDataSet(gse, design = ~ cellLine + dexamethasone)
```
