---
title: "DESeq2 Differential Expression Analysis"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

This weeks Lessons were adapted from
<https://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html>

This week, we're going to do things a little backwards.

In Lesson 19, we'll go over the DESeq2 Differential Expression Analysis
analysis, and then in Lesson 20 we'll go back and look at how to perform
some Sample QC. If you're doing Differential Expression analysis, it's
recommended that you perform the QC prior to Differential Expression
Analysis.

The reasoning behind doing it this way, is that there is a lot of new
libraries, functions, and material so I want to cover the DESeq2
Analysis first before adding in a lot of additional material - sample
QC, Gene Clustering, Salmon, tximeta.

# Differential Expression

This is an introduction to RNAseq Differential Expression Analysis
involving reading in quantitated gene expression data from an RNA-seq
experiment, exploring the data using base R functions and then analysis
with the DESeq2 package.

This lesson assumes a basic familiarity with R, data frames, and
manipulating data with dplyr and \|\>.

The Bioinformatics libraries for R are found in
[Bioconductor](https://www.bioconductor.org/). To install packages from
Bioconductor, you'll need to install the `BiocManager` package. See the
next code section for an example

### Installing Packages from Bioconductor

Please note that there are a lot of packages, that'll need to be
installed for this lesson. It's recommended to use the Class RStudio
Server for this task, as they have already been installed.

```{r eval=FALSE}
install.packages('BiocManager')
BiocManager::install("DESeq2")
library(DESeq2)
```

If you want to install them on your own computer or laptop, you may need
to install additional Linux Packages, as well.

## Additional Reading:

1.  [Conesa et al. A survey of best practices for RNA-seq data analysis.
    *Genome Biology* 17:13
    (2016)](http://genomebiology.biomedcentral.com/articles/10.1186/s13059-016-0881-8).
2.  [Soneson et al. "Differential analyses for RNA-seq: transcript-level
    estimates improve gene-level inferences." *F1000Research* 4
    (2015)](https://f1000research.com/articles/4-1521/v2).
3.  Abstract and introduction sections of [Himes et al. "RNA-Seq
    transcriptome profiling identifies CRISPLD2 as a glucocorticoid
    responsive gene that modulates cytokine function in airway smooth
    muscle cells." *PLoS ONE* 9.6 (2014):
    e99625](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0099625).
4.  Review the [*Introduction*
    (10.1)](http://r4ds.had.co.nz/tibbles.html#introduction-4),
    [*Tibbles vs. data.frame*
    (10.3)](http://r4ds.had.co.nz/tibbles.html#tibbles-vs.data.frame),
    and [*Interacting with Older Code*
    (10.4)](http://r4ds.had.co.nz/tibbles.html#interacting-with-older-code)
    sections of the [***R for Data Science***
    **book**](http://r4ds.had.co.nz/tibbles.html). We will initially be
    using the `read_*` functions from the [**readr**
    package](http://readr.tidyverse.org/). These functions load data
    into a *tibble* instead of R's traditional data.frame. Tibbles are
    data frames, but they tweak some older behaviors to make life a
    little easier. These sections explain the few key small differences
    between traditional data.frames and tibbles.

## Exploring the Dataset

We'll load `Tidyverse` to load the *tibbles* and load the sample data.
Remember, the data for this lesson comes from:

> Himes *et al*. "RNA-Seq Transcriptome Profiling Identifies CRISPLD2 as
> a Glucocorticoid Responsive Gene that Modulates Cytokine Function in
> Airway Smooth Muscle Cells." *PLoS ONE*. [2014 Jun
> 13;9(6):e99625](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0099625).
> PMID: [24926665](http://www.ncbi.nlm.nih.gov/pubmed/24926665).

They had 8 total samples - 4 different airway smooth muscle cell lines
which each has a sample that was untreated and treated with
Dexamethasone.

```{r}
library(tidyverse)
# read in sample info
sampleData <- read_csv('sample_info.csv', col_names = TRUE)
sampleData
```

The first column is the NCBI Sequence Read Archive Accession Number for
the sequence sample. Go to [NCBI SRA](https://www.ncbi.nlm.nih.gov/sra)
and search for *SRR1039508* and *SRR1039509*. What information is
provided about the sequence? Does it match what you have in your
*sampleData*

# Importing the Counts Data

```{r}
countsData <- read_csv('counts_data.csv',col_names=TRUE)
countsData
```

The Counts data was generated by aligning the FASTQ reads to the
transcriptome and counting how many reads mapped to each gene. The first
column of *ensgene* is the Ensembl Gene Accession. Goto
[Ensembl](https://useast.ensembl.org/) and search for *ENSG00000000003*,
what gene is the is the accession for? What information is provided
about the gene on Ensembl?

## RNASeq and Low Counts Data

One important fact about RNASeq Data is that a lot of counts may be
close to zero. To look at this, let's use *mutate* to create a new
column called *controlmean* and *treatedmean*.

```{r echo=FALSE, results="markup"}
meancounts <- countsData |> 
  mutate(controlmean=(SRR1039508+SRR1039512+SRR1039516+SRR1039520)/4) |> 
  mutate(treatedmean=(SRR1039509+SRR1039513+SRR1039517+SRR1039521)/4)

meancounts
```

We can plot a histogram or distribution plot of the values.

```{r simplescatter, echo=F}
ggplot(meancounts, aes(controlmean)) + geom_freqpoly(binwidth=100)
```

We see that there are a lot of low counts. We'll need to take care of
this later

We read the Tables in as Tibbles, but DESeq2 requires R *data.frame*
datatypes. Let's convert them to *data.frame* and verify that they are
the correct datatype

```{r}
countsData <- as.data.frame(countsData)
sampleData <- as.data.frame(sampleData)
head(countsData)
head(sampleData)
```

```{r}
class(countsData)
class(sampleData)
```

Are they both *data.frame* datatypes?

## Quick QCing our Tables Order

DESeq expects that the rows of *sampleData* are in the same order as the
*countsData*.

We need to check that verify that our *countsData* columns are in the
same order as our *sampleData* columns. To get the column names of the
*countsData* we can use *names()*.

```{r}
names(countsData)
```

We'll want to exclude the the first column name because this isn't one
of the samples. To exclude the first item in the character vector, we
can use negative indexing.

```{r}
names(countsData)[-1]
```

```{r}
sampleData$sample
```

```{r}
all(names(countsData)[-1]==sampleData$sample)
```

## Constructing the DESeq2 Object

```{r sumexp, echo=FALSE}
par(mar=c(0,0,0,0))
plot(1,1,xlim=c(0,100),ylim=c(0,100),bty="n",
     type="n",xlab="",ylab="",xaxt="n",yaxt="n")
polygon(c(45,90,90,45),c(5,5,70,70),col="pink",border=NA)
polygon(c(45,90,90,45),c(68,68,70,70),col="pink3",border=NA)
text(67.5,40,"assay(s)")
text(67.5,35,'e.g. "counts", ...')
polygon(c(10,40,40,10),c(5,5,70,70),col="skyblue",border=NA)
polygon(c(10,40,40,10),c(68,68,70,70),col="skyblue3",border=NA)
text(25,40,"rowRanges")
polygon(c(45,90,90,45),c(75,75,95,95),col="palegreen",border=NA)
polygon(c(45,47,47,45),c(75,75,95,95),col="palegreen3",border=NA)
text(67.5,85,"colData")
```

```{r eval=FALSE}
library(DESeq2)
dds <- DESeqDataSetFromMatrix(countData = ...,  # Counts Data
                              colData = ...,  # Sample Data
                              design = ~ ..., # Grouping or Treatment
                              tidy=TRUE   # If First column of Counts Data is the Row Names for Counts
                              )
```

For the *design*, we need to specify how we want to group or define the
treatment. For a single treatment or group, we specify the column *\~
columnName*. In our case, it's going to be `~ dexamethasone`. If you
have multiple treatments or groupings, you can add them using the `+`
sign. For example if we wanted to also split by *cellLine*, we'd define
design as ``` ~``cellLine +``dexamethasone ```

```{r}
library(DESeq2)
dds <- DESeqDataSetFromMatrix(countData = countsData,
                              colData = sampleData,
                              design = ~ cellLine + dexamethasone,
                              tidy=TRUE
                              )

dds
```

If the columns, supplied to the design argument aren't *factors*,
they'll be converted to *factors*. The `rownames(...)` describes the
number and names used for the rows.

How many genes are there represented in this dataset?

What type of an object is the DESeq2 object?

```{r}
class(dds)
typeof(dds)
```

We see that the class is a *DESeqDataSet* and the type is an *S4*. An
*S4* object is a new type of object that we haven't seen before. It's
actually covered in an advanced section of R programming, when you want
to create your own *class/type* of object that also contains your own
set of functions called *methods*. We don't need to know everything
about it. If you click on the variable in the *RStudio Environment
Windows*, you can explore it a bit more.

**More on S4:**

-   [S4 Geeks for Geeks](https://www.geeksforgeeks.org/r-s4-class/)

-   [S4: Advanced R](https://adv-r.hadley.nz/s4.html)

Notice that it stores the Sample Data, Counts Data, and Design.

## Filtering out Low Counts

Recall, the histogram above. There were a lot of low counts values.
Let's remove calls that are lower than 10 with

Our count matrix contains many rows with only zeros, and additionally
many rows with only a few fragments total. In order to reduce the size
of the object, and to increase the speed of our functions, we can remove
the rows that have no or nearly no information about the amount of gene
expression.

Here we perform pre-filtering to keep only rows that have a count of at
least 10 for a minimal number of samples. The count of 10 is a
reasonable choice for bulk RNA-seq. A recommendation for the minimal
number of samples is to specify the smallest group size, e.g. here there
are 4 samples in each group. If there are not discrete groups, one can
use the minimal number of samples where non-zero counts would be
considered interesting.

```{r}
nrow(dds)
#smallestGroupSize <- 4
keep <- rowSums(counts(dds) >= 10) #>= smallestGroupSize
dds <- dds[keep,]
nrow(dds)
```

We now see that we've reduced the number of rows(ie genes) from 64,102
to 16,139

## Ensuring our Reference Group is Properly Set

**Note:** it is prefered in R that the first level of a factor be the
reference level (e.g. control, or untreated samples). In this case, when
the `sampleData` table was assembled the treated samples were set as the
reference, but we want the untreated as reference. Therefore, we need to
use *relevel* as shown below.

*relevel* is a function, which decides how the variables will be coded,
and how contrasts will be computed. For a two-group comparison, the use
of *relevel* to change the reference level would flip the sign of a
coefficient associated with a contrast between the two groups.

```{r}
dds$dexamethasone <- relevel(dds$dexamethasone, ref = "untreated")
dds$dexamethasone
```

## Running DESeq Differential Expression Analysis

Once we've ensured our counts are filtered and levels are in the correct
order, we are ready to run DESeq2 analysis.

To run the analysis, we need to pass the *DESeqDataset* to the `DESeq()`
function. This will estimate the size factors, the dispersion, fit a
model to the counts and finally perform the hypothesis tests. The
results are stored back in the *DESeqDataset*.

```{r}
dds <- DESeq(dds)
```

To get the results table, we need to use the `results()` function to
extracts a result table from the DESeq analysis giving columns for the
base means across samples, log2 fold changes, standard errors, test
statistics, p-value, and adjusted p-value.

```{r}
res <- results(dds)

res
```

The first column, `baseMean`, is a just the average of the normalized
count values, divided by the size factors, taken over all samples in the
*DESeqDataSet*. The remaining four columns refer to a specific contrast,
namely the comparison of the `treated` level over the `untreated` level
for the factor variable `dexamethasone`.

The column `log2FoldChange` is the effect size estimate. It tells us how
much the gene's expression seems to have changed due to treatment with
dexamethasone in comparison to untreated samples. This value is reported
on a logarithmic scale to base 2: for example, a log2 fold change of 1.5
means that the gene's expression is increased by a multiplicative factor
of $2^{1.5} \approx 2.82$.

Of course, this estimate has an uncertainty associated with it, which is
available in the column `lfcSE`, the standard error estimate for the
log2 fold change estimate. We can also express the uncertainty of a
particular effect size estimate as the result of a statistical test. The
purpose of a test for differential expression is to test whether the
data provides sufficient evidence to conclude that this value is really
different from zero. *DESeq2* performs for each gene a *hypothesis test*
to see whether evidence is sufficient to decide against the *null
hypothesis* that there is zero effect of the treatment on the gene and
that the observed difference between treatment and control was merely
caused by experimental variability (i.e., the type of variability that
you can expect between different samples in the same treatment group).
As usual in statistics, the result of this test is reported as a *p*
value, and it is found in the column `pvalue`. Remember that a *p* value
indicates the probability that a fold change as strong as the observed
one, or even stronger, would be seen under the situation described by
the null hypothesis.

We can also summarize the results with the following line of code, which
reports some additional information, that will be covered in later
sections.

## Exploring the Results of DESeq2 Analysis

```{r}
summary(res)
```

Notice that the summary provides some information about the adjusted
P-value cutoff. By default, independent filtering wasn't performed, so
the False Discovery Rate (FDR) alpha value for the P-value adjustment
was set to 0.1.

Note that there are many genes with differential expression due to
dexamethasone treatment at the FDR level of 10%. This makes sense, as
the smooth muscle cells of the airway are known to react to
glucocorticoid steroids. However, there are two ways to be more strict
about which set of genes are considered significant:

-   lower the false discovery rate threshold (the threshold on `padj` in
    the results table)
-   raise the log2 fold change threshold from 0 using the `lfcThreshold`
    argument of *results*

If we lower the false discovery rate threshold, we should also inform
the `results()` function about it, so that the function can use this
threshold for the optimal independent filtering that it performs:

To change the FDR alpha p-value cutoff value, you can use the
*results()* function to specify an *alpha* new cutoff value.

```{r}
res0.01 <- results(dds, alpha = 0.01)
summary(res0.01)

```

If we want to raise the log2 fold change threshold, so that we test for
genes that show more substantial changes due to treatment, we simply
supply a value on the log2 scale. For example, by specifying
`lfcThreshold = 1`, we test for genes that show significant effects of
treatment on gene counts more than doubling or less than halving,
because $2^1 = 2$.

```{r}
resLFC1 <- results(dds, lfcThreshold=1)
table(resLFC1$padj < 0.1)
```

## Multiple testing

In high-throughput biology, we are careful to not use the *p* values
directly as evidence against the null, but to correct for *multiple
testing*.

What would happen if we were to simply threshold the *p* values at a low
value, say 0.05?

There are `r sum(res$pvalue < .05, na.rm=TRUE)` genes with a *p* value
below 0.05 among the `r sum(!is.na(res$pvalue))` genes for which the
test succeeded in reporting a *p* value:

```{r sumres}
sum(res$pvalue < 0.05, na.rm=TRUE)
sum(!is.na(res$pvalue))
```

Now, assume for a moment that the null hypothesis is true for all genes,
i.e., no gene is affected by the treatment with dexamethasone. Then, by
the definition of the *p* value, we expect up to 5% of the genes to have
a *p* value below 0.05. This amounts to
`r round(sum(!is.na(res$pvalue)) * .05 )` genes. If we just considered
the list of genes with a *p* value below 0.05 as differentially
expressed, this list should therefore be expected to contain up to
`r round(sum(!is.na(res$pvalue)) * .05)` /
`r sum(res$pvalue < .05, na.rm=TRUE)` =
`r round(sum(!is.na(res$pvalue))*.05 / sum(res$pvalue < .05, na.rm=TRUE) * 100)`%
false positives.

```{r}
sum(!is.na(res$pvalue))*.05 / sum(res$pvalue < .05, na.rm=TRUE) * 100
```

*DESeq2* uses the Benjamini-Hochberg (BH) adjustment
[@Benjamini1995Controlling] as implemented in the base R *p.adjust*
function; in brief, this method calculates for each gene an adjusted *p*
value that answers the following question: if one called significant all
genes with an adjusted *p* value less than or equal to this gene's
adjusted *p* value threshold, what would be the fraction of false
positives (the *false discovery rate*, FDR) among them, in the sense of
the calculation outlined above? These values, called the BH-adjusted *p*
values, are given in the column `padj` of the `res` object.

The FDR is a useful statistic for many high-throughput experiments, as
we are often interested in reporting or focusing on a set of interesting
genes, and we would like to put an upper bound on the percent of false
positives in this set.

Hence, if we consider a fraction of 10% false positives acceptable, we
can consider all genes with an adjusted *p* value below 10% = 0.1 as
significant. How many such genes are there?

```{r}
sum(res$padj < 0.1, na.rm=TRUE)
```

We subset the results table to these genes and then sort it by the log2
fold change estimate to get the significant genes with the strongest
down-regulation:

```{r}
resSig <- subset(res, padj < 0.1)
head(resSig[ order(resSig$log2FoldChange), ])
```

...and with the strongest up-regulation:

```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```

------------------------------------------------------------------------

# Plotting results

## Counts plot

A quick way to visualize the counts for a particular gene is to use the
*plotCounts* function that takes as arguments the *DESeqDataSet*, a gene
name, and the group over which to plot the counts (figure below).

```{r plotcounts}
topGene <- rownames(res)[which.min(res$padj)]
plotCounts(dds, gene = topGene, intgroup=c("dexamethasone"))
```

**Normalized counts for a single gene over treatment group.**

We can also make custom plots using the *ggplot* function from the
`r CRANpkg("ggplot2")` package (figures below).

```{r ggplotcountsjitter, fig.width = 4, fig.height = 3}
library("ggbeeswarm")
geneCounts <- plotCounts(dds, gene = topGene, intgroup = c("dexamethasone","cellLine"),
                         returnData = TRUE)
ggplot(geneCounts, aes(x = dexamethasone, y = count, color = cellLine)) +
  scale_y_log10() +  geom_beeswarm(cex = 3)
```

```{r ggplotcountsgroup, fig.width = 4, fig.height = 3}
ggplot(geneCounts, aes(x = dexamethasone, y = count, color = cellLine, group = cellLine)) +
  scale_y_log10() + geom_point(size = 3) + geom_line()
```

**Normalized counts with lines connecting cell lines.** Note that the
*DESeq* test actually takes into account the cell line effect, so this
figure more closely depicts the difference being tested.

## MA-plot

An *MA-plot* [@Dudoit2002Statistical] provides a useful overview for the
distribution of the estimated coefficients in the model, e.g. the
comparisons of interest, across all genes. On the y-axis, the "M" stands
for "minus" -- subtraction of log values is equivalent to the log of the
ratio -- and on the x-axis, the "A" stands for "average". You may hear
this plot also referred to as a mean-difference plot, or a Bland-Altman
plot.

Before making the MA-plot, we use the *lfcShrink* function to shrink the
log2 fold changes for the comparison of dex treated vs untreated
samples. There are three types of shrinkage estimators in *DESeq2*,
which are covered in the [DESeq2
vignette](https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html).
Here we specify the *apeglm* method for shrinking coefficients, which is
good for shrinking the noisy LFC estimates while giving low bias LFC
estimates for true large differences [@Zhu2018]. To use *apeglm* we
specify a coefficient from the model to shrink, either by name or number
as the coefficient appears in `resultsNames(dds)`.

```{r plotma}
library("apeglm")
resultsNames(dds)
res <- lfcShrink(dds, coef="cellLine_N061011_vs_N052611", type="apeglm")
plotMA(res, ylim = c(-5, 5))
```

```{r}
res
```

If it is necessary to specify a contrast not represented in
`resultsNames(dds)`, either of the other two shrinkage methods can be
used, or in some cases, re-factoring the relevant variables and running
`nbinomWaldTest` followed by `lfcShrink` is sufficient. See the DESeq2
vignette for more details.

**An MA-plot of changes induced by treatment.** The log2 fold change for
a particular comparison is plotted on the y-axis and the average of the
counts normalized by size factor is shown on the x-axis. Each gene is
represented with a dot. Genes with an adjusted *p* value below a
threshold (here 0.1, the default) are shown in red.

The *DESeq2* package uses a Bayesian procedure to moderate (or "shrink")
log2 fold changes from genes with very low counts and highly variable
counts, as can be seen by the narrowing of the vertical spread of points
on the left side of the MA-plot. As shown above, the *lfcShrink*
function performs this operation. For a detailed explanation of the
rationale of moderated fold changes, please see the *DESeq2* paper
[@Love2014Moderated].

If we had not used statistical moderation to shrink the noisy log2 fold
changes, we would have instead seen the following plot:

```{r plotmaNoShr}
res.noshr <- results(dds, name="dexamethasone_treated_vs_untreated")
plotMA(res.noshr, ylim = c(-5, 5))
```

We can label individual points on the MA-plot as well. Here we use the
*with* R function to plot a circle and text for a selected row of the
results object. Within the *with* function, only the `baseMean` and
`log2FoldChange` values for the selected rows of `res` are used.

```{r plotmalabel}
plotMA(res, ylim = c(-5,5))
topGene <- rownames(res)[which.min(res$padj)]
with(res[topGene, ], {
  points(baseMean, log2FoldChange, col="dodgerblue", cex=2, lwd=2)
  text(baseMean, log2FoldChange, topGene, pos=2, col="dodgerblue")
})
```

Another useful diagnostic plot is the histogram of the *p* values
(figure below). This plot is best formed by excluding genes with very
small counts, which otherwise generate spikes in the histogram.

```{r histpvalue2}
hist(res$pvalue[res$baseMean > 1], breaks = 0:20/20,
     col = "grey50", border = "white")
```

**Histogram of *p* values for genes with mean normalized count larger
than 1.**

## Volcano Plot

```{r}
#reset par
par(mfrow=c(1,1))
# Make a basic volcano plot
with(res, plot(log2FoldChange, -log10(pvalue), pch=20, main="Volcano plot", xlim=c(-5,5)))

# Add colored points: blue if padj<0.01, red if log2FC>1 and padj<0.05)
with(subset(res, padj<.01 ), points(log2FoldChange, -log10(pvalue), pch=20, col="blue"))
with(subset(res, padj<.01 & abs(log2FoldChange)>2), points(log2FoldChange, -log10(pvalue), pch=20, col="red"))
```

```{r}
library(EnhancedVolcano)
EnhancedVolcano(res,
    lab = rownames(res),
    x = 'log2FoldChange',
    y = 'padj')
```

## Independent filtering

The MA plot highlights an important property of RNA-seq data. For weakly
expressed genes, we have no chance of seeing differential expression,
because the low read counts suffer from such high Poisson noise that any
biological effect is drowned in the uncertainties from the sampling at a
low rate. We can also show this by examining the ratio of small *p*
values (say, less than 0.05) for genes binned by mean normalized count.
We will use the results table subjected to the threshold to show what
this looks like in a case when there are few tests with small *p* value.

In the following code chunk, we create bins using the *quantile*
function, bin the genes by base mean using *cut*, rename the levels of
the bins using the middle point, calculate the ratio of *p* values less
than 0.05 for each bin, and finally plot these ratios (figure below).

```{r sensitivityovermean, fig.width=6}
qs <- c(0, quantile(resLFC1$baseMean[resLFC1$baseMean > 0], 0:6/6))
bins <- cut(resLFC1$baseMean, qs)
levels(bins) <- paste0("~", round(signif((qs[-1] + qs[-length(qs)])/2, 2)))
fractionSig <- tapply(resLFC1$pvalue, bins, function(p)
                          mean(p < .05, na.rm = TRUE))
barplot(fractionSig, xlab = "mean normalized count",
                     ylab = "fraction of small p values")
```

**The ratio of small *p* values for genes binned by mean normalized
count.** The *p* values are from a test of log2 fold change greater than
1 or less than -1. This plot demonstrates that genes with very low mean
count have little or no power, and are best excluded from testing.

At first sight, there may seem to be little benefit in filtering out
these genes. After all, the test found them to be non-significant
anyway. However, these genes have an influence on the multiple testing
adjustment, whose performance improves if such genes are removed. By
removing the low count genes from the input to the FDR procedure, we can
find more genes to be significant among those that we keep, and so
improved the power of our test. This approach is known as *independent
filtering*.

The *DESeq2* software automatically performs independent filtering that
maximizes the number of genes with adjusted *p* value less than a
critical value (by default, `alpha` is set to 0.1). This automatic
independent filtering is performed by, and can be controlled by, the
*results* function.

The term *independent* highlights an important caveat. Such filtering is
permissible only if the statistic that we filter on (here the mean of
normalized counts across all samples) is independent of the actual test
statistic (the *p* value) under the null hypothesis. Otherwise, the
filtering would invalidate the test and consequently the assumptions of
the BH procedure. The independent filtering software used inside
*DESeq2* comes from the `r Biocpkg("genefilter")` package, that contains
a reference to a paper describing the statistical foundation for
independent filtering [@Bourgon2010Independent].

## Annotating and exporting results

Our result table so far only contains the Ensembl gene IDs, but
alternative gene names may be more informative for interpretation.
Bioconductor's annotation packages help with mapping various ID schemes
to each other. We load the `r Biocpkg("AnnotationDbi")` package and the
annotation package `r Biocannopkg("org.Hs.eg.db")`:

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```

This is the organism annotation package ("org") for *Homo sapiens*
("Hs"), organized as an *AnnotationDbi* database package ("db"), using
Entrez Gene IDs ("eg") as primary key. To get a list of all available
key types, use:

```{r}
columns(org.Hs.eg.db)
```

We can use the *mapIds* function to add individual columns to our
results table. We provide the row names of our results table as a key,
and specify that `keytype=ENSEMBL`. The `column` argument tells the
*mapIds* function which information we want, and the `multiVals`
argument tells the function what to do if there are multiple possible
values for a single input value. Here we ask to just give us back the
first one that occurs in the database. To add the gene symbol and Entrez
ID, we call *mapIds* twice.

```{r}
ens.str <- substr(rownames(res), 1, 15)
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=ens.str,
                     column="SYMBOL",
                     keytype="ENSEMBL",
                     multiVals="first")
res$entrez <- mapIds(org.Hs.eg.db,
                     keys=ens.str,
                     column="ENTREZID",
                     keytype="ENSEMBL",
                     multiVals="first")
```

Now the results have the desired external gene IDs:

```{r}
resOrdered <- res[order(res$pvalue),]
head(resOrdered)
```

## Exporting results

You can easily save the results table in a CSV file that you can then
share or load with a spreadsheet program such as Excel. The call to
*as.data.frame* is necessary to convert the *DataFrame* object
(`r Biocpkg("IRanges")` package) to a *data.frame* object that can be
processed by *write.csv*. Here, we take just the top 100 genes for
demonstration.

```{r eval=FALSE}
resOrderedDF <- as.data.frame(resOrdered)
write.csv(resOrderedDF, file = "results.csv")
```
