---
title: "Lesson 20 RNASeq Sample QC"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

# Exploratory analysis and visualization

We saw in Lesson 19 that DESeq relies on the raw counts. But if we used
the raw counts in the Sample QC,

```{r}
library(tidyverse)
sampleData <- read_csv('sample_info.csv', col_names=TRUE)
countsData <- read_csv('counts_data.csv',col_names=TRUE)
```

```{r}
countsData = as.data.frame(countsData)
sampleData = as.data.frame(sampleData)
dds <- DESeqDataSetFromMatrix(countData = countsData,
                              colData = sampleData,
                              design = ~ dexamethasone,
                              tidy=TRUE
                             )
```

## Pre-filtering the dataset

Our count matrix with our *DESeqDataSet* contains many rows with only
zeros, and additionally many rows with only a few fragments total. In
order to reduce the size of the object, and to increase the speed of our
functions, we can remove the rows that have no or nearly no information
about the amount of gene expression. Here we perform pre-filtering to
keep only rows that have a count of at least 10 for a minimal number of
samples. The count of 10 is a reasonable choice for bulk RNA-seq. A
recommendation for the minimal number of samples is to specify the
smallest group size, e.g. here there are 4 samples in each group. If
there are not discrete groups, one can use the minimal number of samples
where non-zero counts would be considered interesting. Additional
weighting/filtering to improve power is applied at a later step in the
workflow.

```{r}
nrow(dds)
smallestGroupSize <- 4
keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
dds <- dds[keep,]
nrow(dds)
```

## The variance stabilizing transformation and the rlog

Many common statistical methods for exploratory analysis of
multidimensional data, for example, *clustering* and *principal
componentsanalysis* (PCA), work best for data that generally has the
same range of variance at different ranges of the mean values. When the
expected amount of variance is approximately the same across different
mean values, the data is said to be *homoskedastic*. For RNA-seq counts,
however, the expected variance grows with the mean. For example, if one
performs PCA directly on a matrix of counts or normalized counts (e.g.
correcting for differences in sequencing depth), the resulting plot
typically depends mostly on the genes with *highest* counts because they
show the largest absolute differences between samples.

A simple and often used strategy to avoid this is to take the logarithm
of the normalized count values plus a pseudocount of 1; however,
depending on the choice of pseudocount, now the genes with the very
*lowest* counts will contribute a great deal of noise to the resulting
plot, because taking the logarithm of small counts actually inflates
their variance. We can quickly show this property of counts with some
simulated data (here, Poisson counts with a range of lambda from 0.1 to
100). We plot the standard deviation of each row (genes) against the
mean:

```{r meanSdCts}
lambda <- 10^seq(from = -1, to = 2, length = 1000)
cts <- matrix(rpois(1000*100, lambda), ncol = 100)
library("vsn")
meanSdPlot(cts, ranks = FALSE)
```

And for logarithm-transformed counts:

```{r meanSdLogCts}
log.cts.one <- log2(cts + 1)
meanSdPlot(log.cts.one, ranks = FALSE)
```

The logarithm with a small pseudocount amplifies differences when the
values are close to 0. The low count genes with low signal-to-noise
ratio will overly contribute to sample-sample distances and PCA plots.

As a solution, *DESeq2* offers two transformations for count data that
stabilize the variance across the mean: the *variance stabilizing
transformation* (VST) for negative binomial data with a dispersion-mean
trend [@Anders2010Differential], implemented in the *vst* function, and
the *regularized-logarithm transformation* or *rlog*
[@Love2014Moderated].

For genes with high counts, both the VST and the rlog will give similar
result to the ordinary log2 transformation of normalized counts. For
genes with lower counts, however, the values are shrunken towards a
middle value. The VST or rlog-transformed data then become approximately
homoskedastic (more flat trend in the *meanSdPlot*), and can be used
directly for computing distances between samples, making PCA plots, or
as input to downstream methods which perform best with homoskedastic
data.

**Which transformation to choose?** The VST is much faster to compute
and is less sensitive to high count outliers than the rlog. The rlog
tends to work well on small datasets (n \< 30), potentially
outperforming the VST when there is a wide range of sequencing depth
across samples (an order of magnitude difference). We therefore
recommend the VST for medium-to-large datasets (n \> 30). You can
perform both transformations and compare the `meanSdPlot` or PCA plots
generated, as described below.

Note that the two transformations offered by DESeq2 are provided for
applications *other* than differential testing.

For differential testing, the *DESeq* function applies to raw counts,
which also takes into account the dependence of the variance of counts
on the mean value during the dispersion estimation step.

Both *vst* and *rlog* return a *DESeqTransform* object which is based on
the *SummarizedExperiment* class. The transformed values are no longer
counts, and are stored in the *assay* slot. The *colData* that was
attached to `dds` is still accessible:

```{r vst}
vsd <- vst(dds, blind = FALSE)
head(assay(vsd), 3)
colData(vsd)
```

Again, for the *rlog*:

```{r rlog}
rld <- rlog(dds, blind = FALSE)
head(assay(rld), 3)
```

In the above function calls, we specified `blind = FALSE`, which means
that differences between cell lines and treatment (the variables in the
design) will not contribute to the expected variance-mean trend of the
experiment. The experimental design is not used directly in the
transformation, only in estimating the global amount of variability in
the counts. For a fully *unsupervised* transformation, one can set
`blind = TRUE` (which is the default).

To show the effect of the transformation, in the figure below we plot
the first sample against the second, first simply using the *log2*
function (after adding 1, to avoid taking the log of zero), and then
using the VST and rlog-transformed values. For the *log2* approach, we
need to first estimate *size factors* to account for sequencing depth,
and then specify `normalized=TRUE`. Sequencing depth correction is done
automatically for the *vst* and *rlog*.

```{r transformplot, fig.width = 6, fig.height = 2.5}
library("dplyr")
library("ggplot2")

dds <- estimateSizeFactors(dds)

df <- bind_rows(
  as_data_frame(log2(counts(dds, normalized=TRUE)[, 1:2]+1)) %>%
         mutate(transformation = "log2(x + 1)"),
  as_data_frame(assay(vsd)[, 1:2]) %>% mutate(transformation = "vst"),
  as_data_frame(assay(rld)[, 1:2]) %>% mutate(transformation = "rlog"))
  
colnames(df)[1:2] <- c("x", "y")  

lvls <- c("log2(x + 1)", "vst", "rlog")
df$transformation <- factor(df$transformation, levels=lvls)

ggplot(df, aes(x = x, y = y)) + geom_hex(bins = 80) +
  coord_fixed() + facet_grid( . ~ transformation)  
```

**Scatterplot of transformed counts from two samples**. Shown are
scatterplots using the log2 transform of normalized counts (left), using
the VST (middle), and using the rlog (right). While the rlog is on
roughly the same scale as the log2 counts, the VST has a upward shift
for the smaller values. It is the differences between samples (deviation
from y=x in these scatterplots) which will contribute to the distance
calculations and the PCA plot.

We can see how genes with low counts (bottom left-hand corner) seem to
be excessively variable on the ordinary logarithmic scale, while the VST
and rlog compress differences for the low count genes for which the data
provide little information about differential expression.

## Sample distances

A useful first step in an RNA-seq analysis is often to assess overall
similarity between samples:

-   Which samples are similar to each other?
-   Which are different?
-   Does this fit to the expectation from the experiment's design?

We use the R function *dist* to calculate the Euclidean distance between
samples. To ensure we have a roughly equal contribution from all genes,
we use it on the VST data. We need to transpose the matrix of values
using *t*, because the *dist* function expects the different samples to
be rows of its argument, and different dimensions (here, genes) to be
columns.

```{r}
sampleDists <- dist(t(assay(vsd)))
sampleDists
```

We visualize the distances in a heatmap in a figure below, using the
function *pheatmap* from the `r CRANpkg("pheatmap")` package.

```{r}
library("pheatmap")
library("RColorBrewer")
```

In order to plot the sample distance matrix with the rows/columns
arranged by the distances in our distance matrix, we manually provide
`sampleDists` to the `clustering_distance` argument of the *pheatmap*
function. Otherwise the *pheatmap* function would assume that the matrix
contains the data values themselves, and would calculate distances
between the rows/columns of the distance matrix, which is not desired.
We also manually specify a blue color palette using the
*colorRampPalette* function from the `r CRANpkg("RColorBrewer")`
package.

```{r distheatmap, fig.width = 6.1, fig.height = 4.5}
sampleDistMatrix <- as.matrix( sampleDists )
rownames(sampleDistMatrix) <- paste( vsd$dexamethasone, vsd$cellLine, sep = " - " )
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)
```

**Heatmap of sample-to-sample distances using the variance stabilizing
transformed values.**

Note that we have changed the row names of the distance matrix to
contain treatment type and patient number instead of sample ID, so that
we have all this information in view when looking at the heatmap.

Another option for calculating sample distances is to use the *Poisson
Distance* [@Witten2011Classification], implemented in the
`r CRANpkg("PoiClaClu")` package. This measure of dissimilarity between
counts also takes the inherent variance structure of counts into
consideration when calculating the distances between samples. The
*PoissonDistance* function takes the original count matrix (not
normalized) with samples as rows instead of columns, so we need to
transpose the counts in `dds`.

```{r}
library("PoiClaClu")
poisd <- PoissonDistance(t(counts(dds)))
```

We plot the heatmap in a Figure below.

```{r poisdistheatmap, fig.width = 6.1, fig.height = 4.5}
samplePoisDistMatrix <- as.matrix( poisd$dd )
rownames(samplePoisDistMatrix) <- paste( dds$dexamethasone, dds$cellLine, sep=" - " )
colnames(samplePoisDistMatrix) <- NULL
pheatmap(samplePoisDistMatrix,
         clustering_distance_rows = poisd$dd,
         clustering_distance_cols = poisd$dd,
         col = colors)
```

**Heatmap of sample-to-sample distances using the *Poisson Distance*.**

## PCA plot

Another way to visualize sample-to-sample distances is a principal
components analysis (PCA). In this ordination method, the data points
(here, the samples) are projected onto the 2D plane such that they
spread out in the two directions that explain most of the differences
(figure below). The x-axis is the direction that separates the data
points the most. The values of the samples in this direction are written
*PC1*. The y-axis is a direction (it must be *orthogonal* to the first
direction) that separates the data the second most. The values of the
samples in this direction are written *PC2*. The percent of the total
variance that is contained in the direction is printed in the axis
label. Note that these percentages do not add to 100%, because there are
more dimensions that contain the remaining variance (although each of
these remaining dimensions will explain less than the two that we see).

```{r plotpca, fig.width=6, fig.height=4.5}
plotPCA(vsd, intgroup = c("dexamethasone", "cellLine"))
```

**PCA plot using the VST data.** Each unique combination of treatment
and cell line is given its own color.

Here, we have used the function *plotPCA* that comes with *DESeq2*. The
two terms specified by `intgroup` are the interesting groups for
labeling the samples; they tell the function to use them to choose
colors. We can also build the PCA plot from scratch using the
`r CRANpkg("ggplot2")` package [@Wickham2009Ggplot2]. This is done by
asking the *plotPCA* function to return the data used for plotting
rather than building the plot. See the *ggplot2*
[documentation](http://docs.ggplot2.org/current/) for more details on
using *ggplot*.

```{r}
pcaData <- plotPCA(vsd, intgroup = c( "dexamethasone", "cellLine"), returnData = TRUE)
pcaData
percentVar <- round(100 * attr(pcaData, "percentVar"))
```

We can then use these data to build up a second plot in a figure below,
specifying that the color of the points should reflect dexamethasone
treatment and the shape should reflect the cell line.

```{r ggplotpca, fig.width=6, fig.height=4.5}
ggplot(pcaData, aes(x = PC1, y = PC2, color = dexamethasone, shape = cellLine)) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  ggtitle("PCA with VST data")
```

**PCA plot using the VST values with custom *ggplot2* code.** Here we
specify cell line (plotting symbol) and dexamethasone treatment (color).

From the PCA plot, we see that the differences between cells (the
different plotting shapes) are considerable, though not stronger than
the differences due to treatment with dexamethasone (red vs blue color).
This shows why it will be important to account for this in differential
testing by using a paired design ("paired", because each dex treated
sample is paired with one untreated sample from the *same* cell line).
We are already set up for this design by assigning the formula
`~ cell + dex` earlier.

## PCA plot using Generalized PCA

Another technique for performing dimension reduction on data that is not
Normally distributed (e.g. over-dispersed count data) is *generalized
principal component analysis*, or GLM-PCA, [@Townes2019] as implemented
in the CRAN package `r CRANpkg("glmpca")`. This package takes as input
the count matrix, as well as the number of latent dimensions to fit
(here, we specify 2). As stated by @Townes2019:

> ...we propose the use of GLM-PCA, a generalization of PCA to
> exponential family likelihoods. GLM-PCA operates on raw counts,
> avoiding the pitfalls of normalization. We also demonstrate that
> applying PCA to deviance or Pearson residuals provides a useful and
> fast approximation to GLM-PCA.

```{r}
library("glmpca")
gpca <- glmpca(counts(dds), L=2)
gpca.dat <- gpca$factors
gpca.dat$dexamethasone <- dds$dexamethasone
gpca.dat$cellLine <- dds$cellLine
```

```{r glmpca, fig.width=6, fig.height=4.5}
ggplot(gpca.dat, aes(x = dim1, y = dim2, color = dexamethasone, shape = cellLine)) +
  geom_point(size =3) + coord_fixed() + ggtitle("glmpca - Generalized PCA")
```

## MDS plot

Another plot, very similar to the PCA plot, can be made using the
*multidimensional scaling* (MDS) function in base R. This is useful when
we don't have a matrix of data, but only a matrix of distances. Here we
compute the MDS for the distances calculated from the *VST* data and
plot these in a figure below.

```{r mdsvst, fig.width=6, fig.height=4.5}
mds <- as.data.frame(colData(vsd))  %>%
         cbind(cmdscale(sampleDistMatrix))
ggplot(mds, aes(x = `1`, y = `2`, color = dexamethasone, shape = cellLine)) +
  geom_point(size = 3) + coord_fixed() + ggtitle("MDS with VST data")
```

**MDS plot using VST data.**

In a figure below we show the same plot for the *PoissonDistance*:

```{r mdspois, fig.width=6, fig.height=4.5}
mdsPois <- as.data.frame(colData(dds)) %>%
   cbind(cmdscale(samplePoisDistMatrix))
ggplot(mdsPois, aes(x = `1`, y = `2`, color = dexamethasone, shape = cellLine)) +
  geom_point(size = 3) + coord_fixed() + ggtitle("MDS with PoissonDistances")
```

**MDS plot using the *Poisson Distance*.**

----\

## Gene clustering

Using the Results calculated in Lesson 19

In the sample distance heatmap made previously, the dendrogram at the
side shows us a hierarchical clustering of the samples. Such a
clustering can also be performed for the genes. Since the clustering is
only relevant for genes that actually carry a signal, one usually would
only cluster a subset of the most highly variable genes. Here, for
demonstration, let us select the 20 genes with the highest variance
across samples. We will work with the VST data.

```{r}
library("genefilter")
topVarGenes <- head(order(rowVars(assay(vsd)), decreasing = TRUE), 20)
```

The heatmap becomes more interesting if we do not look at absolute
expression strength but rather at the amount by which each gene deviates
in a specific sample from the gene's average across all samples. Hence,
we center each genes' values across samples, and plot a heatmap (figure
below). We provide a *data.frame* that instructs the *pheatmap* function
how to label the columns.

```{r genescluster}
mat  <- assay(vsd)[ topVarGenes, ]
mat  <- mat - rowMeans(mat)
anno <- as.data.frame(colData(vsd)[, c("cellLine","dexamethasone")])
pheatmap(mat, annotation_col = anno)
```

**Heatmap of relative VST-transformed values across samples.** Treatment
status and cell line information are shown with colored bars at the top
of the heatmap. Blocks of genes that covary across patients. Note that a
set of genes in the heatmap are separating the N061011 cell line from
the others, and there is another set of genes for which the
dexamethasone treated samples have higher gene expression.
